---
title: "Hands-on Exercise 3b: Programming Animated Statistical Graphics with R"
date: "04/28/2026"
date-modified: last-modified
format: 
  html:
    theme: lumen
    toc: true
    toc-depth: 3
    number-sections: false
editor: visual
execute: 
  eval: true
  echo: true
  warning: false
  message: false
  freeze: true
---

## 1 Overview

When telling a visually-driven data story, animated graphics tend to attract the interest of the audience and make deeper impression than static graphics. In this hands-on exercise, we will learn how to create animated data visualisation by using **gganimate** and **plotly r** packages. At the same time, we will also learn how to:

(i) reshape data by using **tidyr** package, and

(ii) process, wrangle and transform data by using **dplyr** package.

### 1.1 Basic Concepts of animation
When creating animations, the plot does not actually move. Instead, many individual plots are built and then stitched together as movie frames, just like an old-school flip book or cartoon.

Each frame is a different plot when conveying motion, which is built using some relevant subset of the aggregate data. The subset drives the flow of the animation when stitched back together.

### 1.2 Terminology

Before we dive into the steps for creating an animated statistical graph, it’s important to understand some of the key concepts and terminology related to this type of visualization.

1. **Frame**: In an animated line graph, each frame represents a different point in time or a different category. When the frame changes, the data points on the graph are updated to reflect the new data.

2. **Animation Attributes**: The animation attributes are the settings that control how the animation behaves. For example, you can specify the duration of each frame, the easing function used to transition between frames, and whether to start the animation from the current frame or from the beginning.

## 2 Getting Started


### (a) Installing and loading libraries

We will use p_load from pacman package to check, install and load the following R packages:

| Package | Description |
|:---|:---|
| [plotly](https://plotly.com/r/) | R library for plotting interactive statistical graphs. |
| [gganimate](https://gganimate.com/) | An ggplot extension for creating animated statistical graphs. |
| [gifski](https://cran.r-project.org/web/packages/gifski/) | Converts video frames to GIF animations. |
| [gapminder](https://cran.r-project.org/web/packages/gapminder/) | An excerpt of the data available at Gapminder.org. |
| [tidyverse](https://www.tidyverse.org/) | A family of modern R packages for data science. |
| [ggpp](https://docs.r4photobiology.info/ggpp/) | Supports data labels and annotations for ggplot2. |

: {tbl-colwidths="[25,75]"}

```{r}
pacman::p_load(readxl, gifski, gapminder, 
               plotly, gganimate, tidyverse,
               ggrepel, ggpp)
```

### (b) Importing Data and Understanding it

:::: {.panel-tabset}
### Importing Data

In this hands-on exercise, the Data worksheet from *GlobalPopulation* Excel workbook will be used.

Write a code chunk to import Data worksheet from GlobalPopulation Excel workbook by using appropriate R package from tidyverse family.

```{r}
col <- c("Country", "Continent")

globalPop <- read_xls("data/GlobalPopulation.xls",
                      sheet="Data") %>%
  mutate(across(all_of(col), as.factor)) %>%
  mutate(Year = as.integer(Year))
```

* [read_xls()](https://readxl.tidyverse.org/reference/read_excel.html) of **readxl** package is used to import the Excel worksheet.

* `mutate(across)` of dplyr package is used to convert all character data type into factor <u>for/across multiple columns</u>.

  * This line applies the `factor()` function to each column specified in the `col` argument. Character to factor. It takes column indices or column names in strings format as inputs, and returns a data frame with new columns for each column in the input data frame, where each new column is the result of applying the specified function to the corresponding column in the input data frame.

  * The `factor(.)` is a way to specify the `factor` works as an argument.

* `mutate` of dplyr package is used to convert data values of Year field into integer.



### Understanding Data

```{r}
glimpse(globalPop)
```
There are 6 columns with numeric (int, dbl) and factor data types.

```{r}
n_distinct(globalPop$Country)
n_distinct(globalPop$Continent)
```
There are 222 unique countries, spanning 6 continents in the dataset.

::::

## 3 Animated Data Visualisation: gganimate methods
[gganimate](https://gganimate.com/) extends the grammar of graphics as implemented by ggplot2 to include the description of animation. It does this by providing a range of new grammar classes that can be added to the plot object in order to customise how it should change with time.

| Arguments | Description |
|:---|:---|
| `transition_*()` | Defines how the data should be spread out and how it relates to itself across time, i.e. you want your data to change. |
| `view_*()` | Defines how the positional scales should change along the animation, i.e. you want your viewpoint to change. |
| `shadow_*()` | Defines how data from other points in time should be presented in the given point in time, i.e. you want the animation to have memory. |
| `enter_*()/exit_*()` | Defines how new data should appear and how old data should disappear during the course of the animation. |
| `ease_aes()` | Defines how different aesthetics should be eased during transitions. |

: {tbl-colwidths="[25,75]"}


:::: {.panel-tabset}
### (a) Static bubble plot
In the code chunk below, the basic ggplot2 functions are used to create a static [bubble plot](https://r-graph-gallery.com/bubble-chart.html). A bubble plot is created when a third numeric variable is assigned to `size =` argument inside a ggplot with geom_point.

```{r}
p <- ggplot(data= globalPop,
       aes(x= Old,
           y=Young,
           size= Population,                   
           color=Country)) +
  geom_point(alpha = 0.7,
             show.legend = FALSE) +
  scale_color_manual(values=country_colors) +  #<<< 'country_colors' from Gapminder lib
  scale_size(range= c(2,12)) +                 #<<<
  labs(title = "Age Distribution Across Time",
       x = '% Aged',
       y= '% Young')+
  theme_bw()

p
```

* The `scale_size(range= c(2,12))` sets the range of point sizes to be used in the plot to between 2 and 12.

* `Population` is mapped to `size` aes in `ggplot`, thus this range parameter controls the min and max size of the points.

* `country_colors` is a built-in named vector from the gapminder package. When we load gapminder, it automatically makes `country_colors` available.

### (b) Animated bubble plot
Now we can incorporate the animation into a basic ggplot graph. Take note that when we run this code, the visual takes some time to render. This is because RStudio is creating a gif representation of our animated chart.

The individual tabs illustrate what each code segment does, and how they evolve when we replace certain sections with other functions.


::: {.panel-tabset}

#### (1) Basic
In the code chunk below:

* [transition_time()](https://gganimate.com/reference/transition_time.html) of **gganimate** is used to create transition through a **continuous variable**, where there are distinct states in time (i.e. Year).

* [ease_aes()](https://gganimate.com/reference/ease_aes.html) is used to control easing of aesthetics. The default is `linear`. Other methods are: quadratic, cubic, quartic, quintic, sine, circular, exponential, elastic, back, and bounce.

* The data points 'move' corresponding to the transition of each country through time. 
```{r}
p +
  transition_time(Year) +
  ease_aes("linear")
```

#### (2) geom_text()

* Text geoms are useful for labeling plots. They can be used by themselves as scatterplots or in combination with other geoms, for example, labeling points or for annotating the height of bars. * [geom_text()](https://ggplot2.tidyverse.org/reference/geom_text.html) adds only text to the plot. * [geom_label()](https://ggplot2.tidyverse.org/reference/geom_text.html) draws a rectangle behind the text, making it easier to read.

```{r}
pt <- ggplot(data= globalPop,
       aes(x= Old,
           y=Young,
           size= Population,
           color=Country)) +
  geom_point(alpha = 0.7,
             show.legend = FALSE) +
    scale_color_manual(values=country_colors) +  #<<< 'country_colors' from Gapminder lib
    scale_size(range= c(2,12)) +
  labs(title = "Age Distribution Across Time",
      subtitle ='Year: {frame_time}',
         x = '% Aged',
         y= '% Young')+
    geom_text(data=globalPop,                #<<<
            aes(x=Old + 1.2,                 #<<< Add/minus a bit to or from the x mapping to shift labels left/right
                y=Young + 1.2,               #<<< Add/minus a bit to or from the y mapping to shift labels up/down  
                label=Country,
                color = "Black",             #<<< Some Country colours were too light when color = Country was used
                hjust=0,
                vjust= 1.2),
                size=2.5,
                show.legend = FALSE)+        #<<< Should be outside of aes() of geom_text()
  theme_bw() +  
  transition_time(Year) +
  ease_aes("linear")

pt
```

#### (3) geom_text_repel()
* Replacing `geom_text()` with `geom_text_repel()` from our last chapter gives us the following results. 
* Note that each label is repelling each other, and not all labels are showing up at the same time.
* These interventions serve to reduce the clutter and improve readability.

```{r}
pr <- ggplot(data= globalPop,
       aes(x= Old,
           y=Young,
           size= Population,
           color=Country)) +
  geom_point(alpha = 0.7,
             show.legend = FALSE) +
    scale_color_manual(values=country_colors) +  #<<< 'country_colors' from Gapminder lib
    scale_size(range= c(2,12)) +
  labs(title = "Age Distribution Across Time",
      subtitle ='Year: {frame_time}',
         x = '% Aged',
         y= '% Young')+
  geom_text_repel(data=globalPop,            #<<<
            aes(x=Old + 1.2,                 #<<< Add/minus a bit to or from the x mapping to shift labels left/right
                y=Young + 1.2,               #<<< Add/minus a bit to or from the y mapping to shift labels up/down 
                label=Country,
                color = "Black",             #<<< Some Country colours were too light when color = Country was used
                hjust=0,
                vjust= 1.2),
            size=2.5,
            show.legend = FALSE,             #<<< Should be outside of aes() of geom_text()
            max.overlaps=)+
  
  theme_bw() +
  transition_time(Year) +
  ease_aes("linear")

pr
```

#### (4) Ring + geom_text_replel()
* We can change the shape of the geom_point to make it resemble a 'ring' look-and-feel. 
* This is done by specifying `shape=21` so that each point has a thick border. We can colour the inside and outside each point separately.
```{r}
ggplot(data= globalPop,
       aes(x= Old,
           y=Young,
           size= Population,
           color=Country,
           label=Country)) +                   #<<< Need to add for geom_text_s to work
  geom_point(alpha = 0.85,
             shape=21,
             stroke =2,
             show.legend = FALSE) +
  scale_color_manual(values=country_colors) +  #<<< 'country_colors' from Gapminder lib
  scale_size(range= c(2,12)) +
  labs(title = "Age Distribution Across Time",
      subtitle ='Year: {frame_time}',
       x = '% Aged',
       y= '% Young')+
  geom_text_repel(data=globalPop,            #<<<
            aes(x=Old + 1.2,                 #<<< Add/minus a bit to or from the x mapping to shift labels left/right
                y=Young + 1.2,               #<<< Add/minus a bit to or from the y mapping to shift labels up/down 
                label=Country,
                color = "Black",             #<<< Some Country colours were too light when color = Country was used
                hjust=0,
                vjust= 1.2),
            size=2.5,
            show.legend = FALSE,             #<<< Should be outside of aes() of geom_text()
            max.overlaps=)+
  theme_bw() +
  transition_time(Year) +
  ease_aes("linear")
```

#### (5) animate both points and text
* We can animate both points and text using `Usinggeom_text_s()` from **ggpp**.
* We can use `anim_save("file_location", plot)` function to export animated chart in GIF format e.g. `anim_save(filename="Images/animation.gif", pr)`
```{r}
ggplot(data= globalPop,
       aes(x= Old,
           y=Young,
           size= Population,
           color=Country,
           label=Country)) +                   #<<< Need to add for geom_text_s to work
  geom_point(alpha = 0.7,
             show.legend = FALSE) +
  scale_color_manual(values=country_colors) +  #<<< 'country_colors' from Gapminder lib
  scale_size(range= c(2,12)) +
  labs(title = "Age Distribution Across Time",
      subtitle ='Year: {frame_time}',
       x = '% Aged',
       y= '% Young')+
  geom_text_s(nudge_x=2, show.legend = FALSE)+
  expand_limits(x=10) +
  theme_bw() +
  transition_time(Year) +
  ease_aes("linear")
```

#### (6) displaying year
Here are some bullet points to describe the code:

* `transition_time(Year)` is used to create transition through a **continuous variable** (Year), where the animation moves through distinct states in time
* `ease_aes("linear")` controls the easing of aesthetics during transitions — `linear` means the animation progresses at a **constant speed** between frames
* `geom_text` with `as.factor(Year)` creates a **large watermark** in the background showing the current year, with `alpha = 0.2` making it semi-transparent and `size = 20` making it large
* `scale_color_manual(values=country_colors)` assigns each country a **consistent colour** from the gapminder palette across all animation frames
* `scale_size(range= c(2,12))` maps the `Population` variable to **bubble size**, with the smallest bubble at size 2 and the largest at size 12
* `geom_text` with `Old + 1.2` and `Young + 1.2` offsets the **country labels** slightly from the data points to avoid overlapping with the bubbles
```{r}
ggplot(data= globalPop,
       aes(x= Old,
           y=Young,
           size= Population,
           color=Country)) +
  geom_point(alpha = 0.7,
             show.legend = FALSE) +
  geom_text(aes(x = max(Old) * 0.5,
                y = max(Young) * 0.5,
                label = as.factor(Year)),
            alpha = 0.2,
            col = "gray",
            size = 20) +
  scale_color_manual(values=country_colors) +
  scale_size(range= c(2,12)) +
  labs(title = "Age Distribution Across Time",
       x = '% Aged',
       y = '% Young') +
  geom_text(data=globalPop,
            aes(x = Old + 1.2,
                y = Young + 1.2,
                label = Country,
                color = "Black"),
            hjust = 0,
            vjust = 1.2,
            size = 2.5,
            show.legend = FALSE) +
  theme_bw() +
  transition_time(Year) +
  ease_aes("linear")
```

#### (7) faceting
* Combining with `facet_wrap()` allows us to arrange the plots in a more space efficient manner when we have a single variable with many levels, such as the 6 unique Continents.
```{r}
p + 
  geom_text_repel(data = globalPop,
                  aes(label = Country),
                  size = 2,
                  max.overlaps = 5,
                  show.legend = FALSE) +
  facet_wrap(~Continent) +
  transition_time(Year)
```

#### (8) moving frame
* Note that `view_follow()` means the **x-axis will pan and zoom** to follow the data as it moves through time, while `fixed_y = TRUE` keeps the **y-axis fixed**. The labels will follow the bubbles as they move!
```{r}
p + 
  geom_text_repel(data = globalPop,
                  aes(label = Country),
                  size = 2,
                  max.overlaps = 5,
                  show.legend = FALSE) +
  transition_time(Year) +
  view_follow(fixed_y = TRUE)
```

#### (9) shadow_wake()

* `shadow_wake(wake_length = 0.1, alpha = FALSE)` creates a **trailing wake effect** behind each bubble as it moves through time — `wake_length = 0.1` means the trail covers 10% of the total animation duration.
* `alpha = FALSE` means the wake trail does **not fade out** in transparency — instead it shrinks in size to create the trailing effect.
* `geom_text_repel()` adds **country labels** to each bubble that automatically reposition themselves to avoid overlapping with other labels and bubbles.
* `max.overlaps = 5` limits the number of overlapping labels shown — labels that would overlap more than 5 others are **hidden** to keep the plot readable.
* `size = 2` keeps the labels **small** so they don't dominate the visualisation.
* `show.legend = FALSE` prevents the text labels from creating an **unnecessary legend entry**.
* Combined with `transition_time(Year)`, the entire plot — bubbles, labels and wake trails — **animates through each year** in the dataset.

```{r}
top_countries <- globalPop %>%
  group_by(Country) %>%
  summarise(avg_pop = mean(Population)) %>%
  slice_max(avg_pop, n = 10) %>%
  pull(Country)

anim <- p + 
  geom_text_repel(data = globalPop %>% filter(Country %in% top_countries),
                  aes(label = Country),
                  color = "black",
                  size = 3,
                  max.overlaps = Inf,
                  show.legend = FALSE,
                  force = 2,
                  segment.color = "grey50",
                  segment.size = 0.3,
                  max.iter = 1000) +
  transition_time(Year) +
  shadow_wake(wake_length = 0.1, alpha = FALSE)

animate(anim, 
        fps = 2,          # frames per second — lower = slower
        duration = 30    # total duration in seconds
        )
```

#### (10) shadow_mark()

* `shadow_mark()` permanently displays **all historical data points** in the background throughout the animation
* `alpha = 0.3` makes background marks **semi-transparent** so current frame stands out
* `size = 0.5` renders background marks **smaller** than current bubbles for visual clarity
* Unlike `shadow_wake()` which only shows recent frames, `shadow_mark()` shows the **complete history** at all times

```{r}
anim_mark <- p + 
  geom_text_repel(data = globalPop %>% filter(Country %in% top_countries),
                  aes(label = Country),
                  color = "black",
                  size = 3,
                  max.overlaps = Inf,
                  show.legend = FALSE,
                  force = 2,
                  segment.color = "grey50",
                  segment.size = 0.3) +
  transition_time(Year) +
  shadow_mark(alpha = 0.3, size = 0.5)

animate(anim_mark, fps = 2, duration = 30)
```
:::
### (c) Animated line graph

* `transition_reveal(Year)` progressively **reveals the line** along the Year dimension, creating a drawing effect
* `geom_point()` adds **data points** along the line to highlight each year's value
* `geom_text_repel()` automatically **repositions labels** to avoid overlapping with lines and points
* `subtitle = "Year: {frame_along}"` dynamically updates the **current year** in the subtitle as the animation progresses
* `scale_color_viridis_d()` assigns each continent a **distinct accessible colour** from the viridis palette
```{r}
globalPop_continent <- globalPop %>%
  group_by(Continent, Year) %>%
  summarise(Old = mean(Old),
            Young = mean(Young),
            Population = sum(Population),
            .groups = "drop")

l <- ggplot(globalPop_continent,
            aes(x = Year, 
                y = Old, 
                group = Continent, 
                color = Continent)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  geom_text_repel(aes(label = Continent),
                  color = "black",
                  size = 3,
                  max.overlaps = Inf,
                  show.legend = FALSE) +
  scale_color_viridis_d() +
  labs(x = "Year", 
       y = "% Aged", 
       title = "% Aged Across Years",
       subtitle = "Year: {frame_along}") +
  theme_bw() +
  transition_reveal(Year)

animate(l, fps = 5, duration = 15)
```

::::

## 4 Animated Data Visualisation: plotly

In **Plotly R** package, both [ggplotly()](https://plotly.com/ggplot2/animations/) and `plot_ly()` support key frame animations through the `frame` argument/aesthetic. They also support an `ids` argument/aesthetic to ensure smooth transitions between objects with the same id (which helps facilitate object constancy).

:::: {.panel-tabset}
### (a) Building an animated bubble plot: ggplotly() method

* The animated bubble plot below includes a play/pause button and a slider component for controlling the animation.
* **Note:** - Appropriate ggplot2 functions are used to create a static bubble plot. The output is then saved as an R object called gg.
* `ggplotly()` is then used to convert the R graphic object into an animated svg object.
* `text = paste0(...)` creates a **custom tooltip** using HTML `<br>` line breaks to display multiple fields cleanly
* `frame = Year` inside `geom_point()` tells ggplotly to **animate by year**
* `ggplotly(gg, tooltip = "text")` uses only the **custom text tooltip** instead of default ggplot2 aesthetics

```{r}
gg <- ggplot(globalPop, 
       aes(x = Old, 
           y = Young, 
           size = Population, 
           colour = Continent,
           text = paste0("Country: ", Country,
                        "<br>Year: ", Year,
                        "<br>Population: ", round(Population, 1),
                        "<br>% Aged: ", Old,
                        "<br>% Young: ", Young))) +
  geom_point(aes(size = Population,
                 frame = Year),
             alpha = 0.7, 
             show.legend = FALSE) +
  scale_size(range = c(2, 12)) +
  labs(x = '% Aged', 
       y = '% Young',
       title = "Age Distribution Across Time") +
  theme_bw() +
  theme(legend.position = "none")

ggplotly(gg, tooltip = "text")
```


### (b)  Building an animated bubble plot: plot_ly() method

* `animation_opts(1000, redraw = FALSE)` sets **1 second per frame** and avoids redrawing the base plot at each transition for smoother animation
* `animation_button()` adds a **play/pause button** positioned at the bottom left
* `sizemode = "diameter"` ensures bubble sizes scale **proportionally** to population values
* `paper_bgcolor` and `plot_bgcolor` set both the **outer and inner** background to white, equivalent to `theme_bw()`

```{r}
bp <- globalPop %>% 
  plot_ly(x = ~Old,
          y = ~Young,
          size = ~Population,
          color = ~Continent,
          frame = ~Year,
          text = ~paste0("Country: ", Country,
                        "<br>Population: ", round(Population, 1),
                        "<br>% Aged: ", Old,
                        "<br>% Young: ", Young),
          hoverinfo = "text",
          type = "scatter",
          mode = "markers",
          marker = list(opacity = 0.7,
                        sizemode = "diameter")) %>%
  layout(title = "Age Distribution Across Time",
         xaxis = list(title = "% Aged",
                      tickcolor = "black",
                      ticklen = 5),
         yaxis = list(title = "% Young",
                      tickcolor = "black",
                      ticklen = 5),
         paper_bgcolor = "white",
         plot_bgcolor = "white") %>%
  animation_opts(1000, redraw = FALSE) %>%
  animation_button(x = 0, xanchor = "left",
                   y = 0, yanchor = "bottom")

bp
```

::::

## 5. References

* Kam, T.S. (2023). [Programming Animated Statistical Graphics with R](https://r4va.netlify.app/chap04).
* [Getting Started](https://gganimate.com/articles/gganimate.html)
* [Building an animation step-by-step with gganimate](https://www.alexcookson.com/post/2020-10-18-building-an-animation-step-by-step-with-gganimate/).
* [Creating a composite gif with multiple gganimate panels](https://solarchemist.se/2021/08/02/composite-gif-gganimate/)