---
title: "Hands-on Exercise 9 -  Information Dashboard Design: R methods"
author: "Andy Tai"
date: "06/22/2026"
date-modified: last-modified
format: 
  html:
    theme: lumen
    toc: true
    toc-depth: 3
    number-sections: false
editor: source
execute: 
  eval: true
  echo: true
  warning: false
  message: false
  freeze: true
---

## 1 Overview

Learning Objectives for this chapter are:

-   create bullet chart by using ggplot2,
-   create sparklines by using ggplot2 ,
-   build industry standard dashboard by using R Shiny.

## 2 Getting Started

:::: {.panel-tabset}

### (a) Install libraries

The following R packages will be used:

-   [tidyverse](https://www.tidyverse.org/) provides a collection of functions for performing data science task such as importing, tidying, wrangling data and visualising data. It is not a single package but a collection of modern R packages including but not limited to [readr](https://readr.tidyverse.org/), [tidyr](https://tidyr.tidyverse.org/), [dplyr](https://dplyr.tidyverse.org/), [ggplot](https://ggplot2.tidyverse.org/), [tibble](https://tibble.tidyverse.org/), [stringr](https://stringr.tidyverse.org/), [forcats](https://forcats.tidyverse.org/) and [purrr](https://purrr.tidyverse.org/).

-   [lubridate](https://r4va.netlify.app/chap31) provides functions to work with dates and times more efficiently.

-   [ggthemes](https://r4va.netlify.app/chap31) is an extension of ggplot2. It provides additional [themes](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/) beyond the basic themes of ggplot2.

-   [gtExtras](https://jthomasmock.github.io/gtExtras/) provides some additional helper functions to assist in creating beautiful tables with [gt](https://gt.rstudio.com/), an R package specially designed for anyone to make wonderful-looking tables using the R programming language.

-   [reactable](https://glin.github.io/reactable/) provides functions to create interactive data tables for R, based on the [React Table](https://github.com/tanstack/table/tree/v7) library and made with [reactR](https://github.com/react-R/reactR).

-   [reactablefmtr](https://kcuilla.github.io/reactablefmtr/) provides various features to streamline and enhance the styling of interactive reactable tables with easy-to-use and highly-customizable functions and themes.

-   svglite

Code chunk below will be used to check if these packages have been installed and also will load them into the working R environment.

```{r}
pacman::p_load(lubridate, ggthemes, reactable,
reactablefmtr, gt, gtExtras, tidyverse,svglite)
```

### (b) Import Data

`odbcConnectAccess()` of [**RODBC**](https://cran.r-project.org/web/packages/RODBC/) package is used used to import a database query table into R.

If an issue arises, [change the R environment from 64bit to 32 bit](https://sites.google.com/site/rforfishandwildlifegrads/home/week-2-databases-types-formats-and-practical-considerations/changing-r-studio-to-default-32-bit-r):

From R Studio \> Tools \> Global Options \> General \> Basic \> R session \> R version: Change \> Use your machine's default version of R (32-bit).

Note: Since R 4.2.0, 32-bit builds are no longer provided.

```{r}

library(RODBC)
con <- odbcConnectAccess2007('data/Coffee Chain.mdb')
coffeechain <- sqlFetch(con, 'CoffeeChain Query')
write_rds(coffeechain, "data/CoffeeChain.rds")
odbcClose(con)
```

### (c) Wrangle Data

The code chunk below is used to import CoffeeChain.rds into R.

```{r}
coffeechain <- read_rds("data/rds/CoffeeChain.rds")
```

::: callout-note
Note: If *coffeechain* is already available in R, above step is optional
:::

The code chunk below is used to aggregate Sales and Budgeted Sales at the Product level.

```{r}
product <- coffeechain %>%
  group_by(`Product`) %>%
  summarise(`target` = sum(`Budget Sales`),
            `current` = sum(`Sales`)) %>%
  ungroup()
```


::::

## 3 Bullet Chart and Sparklines in ggplot2

:::: {.panel-tabset}

## (a) Bullet Chart

The code chunk below is used to plot the bullet charts using ggplot2 functions.

```{r}
ggplot(product, aes(Product, current)) + 
  geom_col(aes(Product, max(target) * 1.01),
           fill="grey85", width=0.85) +
  geom_col(aes(Product, target * 0.75),
           fill="grey60", width=0.85) +
  geom_col(aes(Product, target * 0.5),
           fill="grey50", width=0.85) +
  geom_col(aes(Product, current), 
           width=0.35,
           fill = "black") + 
  geom_errorbar(aes(y = target,
                    x = Product, 
                    ymin = target,
                    ymax= target), 
                width = .4,
                colour = "red",
                size = 1) +
  coord_flip()
```

## (b) Sparklines

We compute the minimum, maximum, end of month sales, as well as the 25 and 75 quantiles.

```{r}
#| fig-width: 12
#| fig-height: 10

sales_report <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  mutate(Month = month(Date)) %>%
  group_by(Month, Product) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup() %>%
  select(Month, Product, Sales)

mins <- group_by(sales_report, Product) %>% 
  slice(which.min(Sales))

maxs <- group_by(sales_report, Product) %>% 
  slice(which.max(Sales))

ends <- group_by(sales_report, Product) %>% 
  filter(Month == max(Month))

quarts <- sales_report %>%
  group_by(Product) %>%
  summarise(quart1 = quantile(Sales, 0.25),
            quart2 = quantile(Sales, 0.75)) %>%
  right_join(sales_report)

ggplot(sales_report, aes(x = Month, y = Sales)) + 
  facet_grid(Product ~ ., scales = "free_y") + 
  geom_ribbon(data = quarts, aes(ymin = quart1, max = quart2), 
              fill = 'grey90') +
  geom_line(size = 0.3) +
  geom_point(data = mins, col = 'red') +
  geom_point(data = maxs, col = 'blue') +
  geom_text(data = mins, aes(label = Sales), vjust = -1, size = 4) +
  geom_text(data = maxs, aes(label = Sales), vjust = 2.5, size = 4) +
  geom_text(data = ends, aes(label = Sales), hjust = 0, nudge_x = 0.5, size = 4) +
  geom_text(data = ends, aes(label = Product), hjust = 0, nudge_x = 2.0, size = 5) +
  expand_limits(x = max(sales_report$Month) + 
                  (0.4 * (max(sales_report$Month) - min(sales_report$Month)))) +
  scale_x_continuous(breaks = seq(1, 12, 1)) +
  scale_y_continuous(expand = c(0.1, 0)) +
  theme_tufte(base_size = 10, base_family = "sans") +
  theme(axis.title = element_blank(), 
        axis.text.y = element_blank(), 
        axis.ticks = element_blank(), 
        strip.text = element_blank())
```
::::

## 4 Static Information Dashboard: gt and gtExtras

Create static information dashboard by using [gt](https://gt.rstudio.com/) and [gtExtras](https://jthomasmock.github.io/gtExtras/) packages.

:::: {.panel-tabset}

### (a) Bullet Chart

```{r}
product %>%
  gt::gt() %>%
  gt_plt_bullet(column = current, 
              target = target, 
              width = 60,
              palette = c("lightblue", 
                          "black")) %>%
  gt_theme_538()
```

### (b) Sparklines

Note that there is a current bug in gtExtras 0.62 version making the background dark even when 'white' is specified. To work around this, I specified 'white' in the first color palette forcing the lines to be 'white' to contrast against the dark background

```{r}
theme_set(theme_minimal())

report <- coffeechain %>%
  mutate(Year = year(Date)) %>%
  filter(Year == "2013") %>%
  mutate(Month = month(Date, 
                       label = TRUE, 
                       abbr = TRUE)) %>%
  group_by(Product, Month) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup()

report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop") %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales',
                   same_limit = FALSE,
                   palette = c("white", "white", "purple", "green", "white")) %>%
  tab_options(table.background.color = "white")
```

### (c) Statistics

```{r}
report %>% 
  group_by(Product) %>% 
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            ) %>%
  gt() %>%
  fmt_number(columns = 4,
    decimals = 2)
```

### (d) table+sparklines

```{r}
spark <- report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop")

sales <- report %>% 
  group_by(Product) %>% 
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T))

sales_data <- left_join(sales, spark)

sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales',
                   same_limit = FALSE,
                   palette = c("white", "white", "purple", "green", "white")) %>%
  tab_options(table.background.color = "white")
```

### (e) table+sparklines+bullet 

```{r}
bullet <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  group_by(Product) %>%
  summarise(Target = sum(`Budget Sales`),
            Actual = sum(Sales)) %>%
  ungroup()

sales_data <- sales_data %>%
  left_join(bullet)

sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales',
                   same_limit = FALSE,
                   palette = c("white", "white", "purple", "green", "white")) %>%
  gt_plt_bullet(column = Actual,
                target = Target,
                width = 28,
                palette = c("lightblue", "black")) %>%
  gt_theme_538()
```

:::: 

## 5 Interactive Information Dashboard: reactable and reactablefmtr

Create interactive information dashboard by using [reactable](https://glin.github.io/reactable/) and [reactablefmtr](https://kcuilla.github.io/reactablefmtr/) packages.

In order to build an interactive sparklines, dataui R package needs to be installed.

```{r}
#| output: false
#| message: false
#| warning: false
remotes::install_github("timelyportfolio/dataui", force = TRUE)
remotes::install_github("kcuilla/reactablefmtr")
```

Load the packages onto R environment

```{r}
#| output: false
#| message: false
#| warning: false
library(dataui)
library(reactablefmtr)
```

:::: {.panel-tabset}

### (a) sparklines

Prepare the list field

```{r}
report <- report %>%
  group_by(Product) %>%
  summarize(`Monthly Sales` = list(Sales))
```

Next, [react_sparkline](https://kcuilla.github.io/reactablefmtr/reference/react_sparkline.html) will be used to plot the sparklines

```{r}
reactable(
  report,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

### (b) change pagesize

By default the pagesize is 10. Argument defaultPageSize is used to change the default setting.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

### (c) points/labels

`highlight_points` argument is used to show the minimum and maximum values points and `label` argument is used to label first and last values.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        labels = c("first", "last")
        )
    )
  )
)
```

### (d) reference line

`statline` argument is used to show the mean line.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        statline = "mean"
        )
    )
  )
)
```

### (e) bandline

bandline can be added by using the `bandline` argument.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        line_width = 1,
        bandline = "innerquartiles",
        bandline_color = "green"
        )
    )
  )
)
```

### (f) sparkbar

Instead of displaying the values as sparklines, we can display them as sparkbars.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkbar(
        report,
        highlight_bars = highlight_bars(
          min = "red", max = "blue"),
        bandline = "innerquartiles",
        statline = "mean")
    )
  )
)
```


::::


