---
title: "Hands-on Exercise 8b: Visualising Geospatial Point Data"
author: "Andy Tai"
date: "06/16/2026"
date-modified: last-modified
format: 
  html:
    theme: lumen
    toc: true
    toc-depth: 3
    number-sections: false
editor: source
execute: 
  eval: true
  echo: true
  warning: false
  message: false
  freeze: true
---

## 1 Overview

Learning Objectives for this Chapter:

-   To import an aspatial data file into R.
-   To convert it into simple point feature data frame and at the same time, to assign an appropriate projection reference to the newly create simple point feature data frame.
-   To plot interactive proportional symbol maps.

## 2 Getting Started

:::: {.panel-tabset}
### (a) Install libraries

The following R packages will be used:

-   Tidyverse:

    -   [**readr**](https://readr.tidyverse.org/) for importing delimited text file,

    -   [**tidyr**](https://tidyr.tidyverse.org/) for tidying data,

    -   [**dplyr**](https://dplyr.tidyverse.org/) for wrangling data and

-   [**sf**](https://cran.r-project.org/web/packages/sf/) for handling geospatial data

-   tmap for plotting choropleth maps

Code chunk below will be used to check if these packages have been installed and also will load them into the working R environment.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

### (b) Import Data

#### The Data

The data set used is called *SGPools_svy21*. The data is in csv file format.

Figure below shows the first 15 records of SGPools_svy21.csv. It consists of seven columns. The XCOORD and YCOORD columns are the x-coordinates and y-coordinates of SingPools outlets and branches. They are in [Singapore SVY21 Projected Coordinates System](https://www.sla.gov.sg/sirent/CoordinateSystems.aspx).

#### Data Import

The code chunk below uses *read_csv()* function of **readr** package to import *SGPools_svy21.csv* into R as a tibble data frame called *sgpools*.

```{r}
sgpools <- read_csv("data/aspatial/SGPools_svy21.csv")
```

### (c) Process Data

#### Examine Data
```{r}
head(sgpools, 2) 
```

::: callout-note
Notice that the *sgpools* data in tibble data frame and not the common R data frame.
:::

#### Creating a sf data frame from an aspatial data frame

The code chunk below converts sgpools data frame into a simple feature data frame by using *st_as_sf()* of **sf** packages

```{r}
sgpools_sf <- st_as_sf(sgpools, 
                       coords = c("XCOORD", "YCOORD"),
                       crs= 3414)
```

::: callout-note
-   The *coords* argument requires the column name of the x-coordinates to be provided first, followed by the column name of the y-coordinates.

-   The *crs* argument requires the coordinates system in epsg format. [EPSG: 3414](https://epsg.io/3414) is Singapore SVY21 Projected Coordinate System to be provided. Country epsg codes can be found at [epsg.io](https://epsg.io/).
:::

Figure below shows the data table of *sgpools_sf*. Notice that a new column called geometry has been added into the data frame.


Display the basic information of the newly created sgpools_sf 

```{r}
head(sgpools_sf, 2)
```

::: callout-note
The output shows that sgppols_sf is in point feature class. It’s epsg ID is 3414. The bbox provides information of the extend of the geospatial data.
:::

:::: 

## 3 Drawing Proportional Symbol Map

:::: {.panel-tabset}

To create an interactive proportional symbol map in R, the view mode of tmap will be used.

```{r}
tmap_mode("view")
```

### (a) Interactive point symbol map

The code chunks below are used to create an interactive point symbol map.

```{r}
tm_shape(sgpools_sf)+
tm_bubbles(col = "red",
           size = 1,
           border.col = "black",
           border.lwd = 1)
```


### (b) Proportional Symbol Map

To draw a proportional symbol map, the numerical variable needs to be assigned to the size visual attribute. The code chunk below shows the variable *Gp1Gp2Winnings* is assigned to size visual attribute.

```{r}
tm_shape(sgpools_sf)+
tm_bubbles(col = "red",
           size = "Gp1Gp2 Winnings",
           border.col = "black",
           border.lwd = 1)
```

::: {.panel-tabset}

#### (i) Colour Visual Attribute

The proportional symbol map can be further improved by using the colour visual attribute. In the code chunk below, *OUTLET_TYPE* variable is used as the colour attribute variable.

```{r}
tm_shape(sgpools_sf)+
tm_bubbles(col = "OUTLET TYPE", 
          size = "Gp1Gp2 Winnings",
          border.col = "black",
          border.lwd = 1)
```


#### (ii) Faceted Plots

**tmap**’s view mode also works with faceted plots. The argument *sync* in *tm_facets()* can be used in this case to produce multiple maps with synchronised zoom and pan settings.

```{r}
tm_shape(sgpools_sf) +
  tm_bubbles(col = "OUTLET TYPE", 
          size = "Gp1Gp2 Winnings",
          border.col = "black",
          border.lwd = 1) +
  tm_facets(by= "OUTLET TYPE",
            nrow = 1,
            sync = TRUE)
```

::: 


Switch **tmap**'s viewer back to plot mode

```{r}
tmap_mode("plot")
```

::::

## 4 References

### All about tmap package

* [tmap: Thematic Maps in R](https://www.jstatsoft.org/article/view/v084i06)
* [tmap](https://cran.r-project.org/web/packages/tmap/index.html)


### Geospatial data wrangling

* [sf: Simple Features for R](https://cran.r-project.org/web/packages/sf/index.html)
* [Simple Features for R: StandardizedSupport for Spatial Vector Data](https://journal.r-project.org/articles/RJ-2018-009/RJ-2018-009.pdf)
* [Reading, Writing and Converting Simple Features](https://cran.r-project.org/web/packages/sf/vignettes/sf2.html)