---
title: "Hands-on Exercise 4b: Visualising Statistical Analysis"
author: "Andy Tai"
date: "05/03/2026"
date-modified: last-modified
format: 
  html:
    theme: lumen
    toc: true
    toc-depth: 3
    number-sections: false
editor: source
execute: 
  eval: true
  echo: true
  warning: false
  message: false
  freeze: true
---

## 1 Learning Outcome
This hands-on exercise focuses on:

* ggstatsplot package to create visual graphics with rich statistical information,
* performance package to visualise model diagnostics, and
* parameters package to visualise model parameters

## 2 Getting Started

:::: {.panel-tabset}
### Installing and loading libraries
* Use the pacman package to check, install and launch the R packages **ggstatplot** and **tidyverse**.
* [ggstatsplot](https://www.indrapatil.com/ggstatsplot/) is an extension of [ggplot2](https://ggplot2.tidyverse.org/) package for creating graphics with details from statistical tests included in the information-rich plots themselves.
```{r}
#| message: false
#| warning: false
pacman::p_load(ggstatsplot, tidyverse, nortest, ggdist)
```


### Importing data
* The code chunk below uses [read_csv()](https://readr.tidyverse.org/reference/read_delim.html) of **readr** package to import Exam_data.csv data file into R and save it as an tibble data frame called `exam_data`.

```{r}
exam <- read_csv("data/Exam_data.csv")
```

### Understanding the data
* The data is a tibble dataframe and contains 322 observations across 7 attributes.
```{r}
glimpse(exam)
```

### Converting data types
```{r}
col <- c("CLASS", "GENDER", "RACE")

exam <- exam %>% 
  mutate(across(col, as.factor)) %>% 
  mutate(ID = as.character(ID))
```

::::

## 3 Various Statistical Tests

| Test | What it Measures | Function |
|:---|:---|:---|
| One-sample | Is this sample's mean/median **different from a known value**? | `gghistostats()` |
| Two-sample | Are these **two groups** different from each other? | `ggbetweenstats()` |
| ANOVA | Are **any of these groups** different from each other? | `ggbetweenstats()` |
| Correlation | Is there a **linear relationship** between two continuous variables? | `ggscatterstats()` |
| Association | Is there a **dependency** between two categorical variables? | `ggbarstats()` |

: {tbl-colwidths="[15,55,30]"}

### 3.1 Compare means/medians
:::: {.panel-tabset}
### (a) 1-sample tests
| Type Argument | Test Used | Nature | When to Use |
|:---|:---|:---|:---|
| `type = "parametric"` | One-sample t-test | Parametric | When data follows normal distribution — tests against **mean** |
| `type = "nonparametric"` | One-sample Wilcoxon signed rank test | Non-parametric | When data does **not** follow normal distribution — tests against **median** |
| `type = "robust"` | One-sample robust t-test | Robust | When data has outliers — uses **trimmed mean** |
| `type = "bayes"` | Bayes Factor test | Bayesian | When you want to quantify evidence **in favour of** null or alternative hypothesis |

: {tbl-colwidths="[20,25,15,40]"}



::: {.panel-tabset}
#### (i) 1-sample t-test
$H_0$: $\mu = 60$ (English scores mean is equal to 60)

$H_1$: $\mu \neq 60$ (English scores mean is not equal to 60)

```{r}
# Parametric (t-test)
set.seed(1234)
gghistostats(
  data = exam,
  x = ENGLISH,
  type = "parametric",
  test.value = 60,
  xlab = "English scores",
  title = "Parametric: One-sample t-test",
  bin.args = list(fill = "seagreen", 
                  color = "white",
                  alpha = 0.8)
) + theme_bw()
```

#### (ii) 1-sample Wilcoxon signed rank test
$H_0$: $\mu = 60$ (English scores mean is equal to 60)

$H_1$: $\mu \neq 60$ (English scores mean is not equal to 60)

```{r}
# Non-parametric (Wilcoxon)
set.seed(1234)
gghistostats(
  data = exam,
  x = ENGLISH,
  type = "nonparametric",
  test.value = 60,
  xlab = "English scores",
  title = "Non-parametric: Wilcoxon signed rank test",
  bin.args = list(fill = "seagreen", 
                  color = "white",
                  alpha = 0.8)
) + theme_bw()
```

#### (iii) 1-sample robust t-test

$H_0$: $\mu = 60$ (English scores mean is equal to 60)

$H_1$: $\mu \neq 60$ (English scores mean is not equal to 60)

```{r}
# Robust
set.seed(1234)
gghistostats(
  data = exam,
  x = ENGLISH,
  type = "robust",
  test.value = 60,
  xlab = "English scores",
  title = "Robust: One-sample robust t-test",
  bin.args = list(fill = "seagreen", 
                  color = "white",
                  alpha = 0.8)
) + theme_bw()
```

#### (iv) Bayes Factor test

* A Bayes factor is the ratio of the likelihood of one particular hypothesis to the likelihood of another. It can be interpreted as a measure of the strength of evidence in favor of one theory among two competing theories.

* That’s because the Bayes factor gives us a way to evaluate the data in favor of a null hypothesis, and to use external information to do so. It tells us what the weight of the evidence is in favor of a given hypothesis.

* When we are comparing two hypotheses: H1 (the alternate hypothesis) and H0 (the null hypothesis), the Bayes Factor is often written as B10, i.e.,
$$
B_{10} = \frac{P(Data | H_1)}{P(Data | H_0)}
$$
* The [Schwarz criterion](https://www.statisticshowto.com/bayesian-information-criterion/) is one of the easiest ways to calculate rough approximation of the Bayes Factor.

| If $B_{10}$ is... | Then you have... |
|:---|:---|
| $> 100$ | Extreme evidence for $H_1$ |
| $30 - 100$ | Very strong evidence for $H_1$ |
| $10 - 30$ | Strong evidence for $H_1$ |
| $3 - 10$ | Moderate evidence for $H_1$ |
| $1 - 3$ | Anecdotal evidence for $H_1$ |
| $1$ | No evidence |
| $1/3 - 1$ | Anecdotal evidence for $H_0$ |
| $1/3 - 1/10$ | Moderate evidence for $H_0$ |
| $1/10 - 1/30$ | Strong evidence for $H_0$ |
| $1/30 - 1/100$ | Very strong evidence for $H_0$ |
| $< 1/100$ | Extreme evidence for $H_0$ |

: {tbl-colwidths="[30,70]"}

$H_0$: $\mu = 60$

$H_1$: $\mu \neq 60$

$$B_{10} = \frac{P(Data | H_1)}{P(Data | H_0)}$$

```{r}
# Bayes
set.seed(1234)
gghistostats(
  data = exam,
  x = ENGLISH,
  type = "bayes",
  test.value = 60,
  xlab = "English scores",
  title = "Bayesian: Bayes Factor test",
  bin.args = list(fill = "seagreen", 
                  color = "white",
                  alpha = 0.8)
) + theme_bw()
```
:::

### (b) 2-sample test

| Type Argument | Test Used | Nature |
|:---|:---|:---|
| `type = "parametric"` | Welch's t-test | Parametric |
| `type = "nonparametric"` or `type = "np"` | Wilcoxon rank-sum test (Mann-Whitney) | Non-parametric |
| `type = "robust"` | Yuen's trimmed means t-test | Robust |
| `type = "bayes"` | Bayesian t-test | Bayesian |

: {tbl-colwidths="[30,45,25]"}

::: {.panel-tabset}
#### (i) Welch's t-test
$H_0$: $\mu_{Female} = \mu_{Male}$ (Mean Math scores of Female and Male are equal)

$H_1$: $\mu_{Female} \neq \mu_{Male}$ (Mean Math scores of Female and Male are not equal)

```{r}
# Parametric (Welch's t-test)
ggbetweenstats(
  data = exam,
  x = GENDER,
  y = MATHS,
  type = "parametric",
  title = "Parametric: Welch's t-test",
  messages = FALSE,
  palette = "Set2",
  package = "RColorBrewer"
) + theme_bw()
```

#### (ii) Wilcoxon rank-sum test

$H_0$: $\mu_{Female} = \mu_{Male}$ (Mean Math scores of Female and Male are equal)

$H_1$: $\mu_{Female} \neq \mu_{Male}$ (Mean Math scores of Female and Male are not equal)

```{r}
# Non-parametric (Wilcoxon rank-sum)
ggbetweenstats(
  data = exam,
  x = GENDER,
  y = MATHS,
  type = "nonparametric",
  title = "Non-parametric: Wilcoxon rank-sum test",
  messages = FALSE,
  palette = "Set2",
  package = "RColorBrewer"
) + theme_bw()
```

#### (iii) Yuen's trimmed means t-test

$H_0$: $\mu_{Female} = \mu_{Male}$ (Mean Math scores of Female and Male are equal)

$H_1$: $\mu_{Female} \neq \mu_{Male}$ (Mean Math scores of Female and Male are not equal)

```{r}
# Robust (Yuen's trimmed means)
ggbetweenstats(
  data = exam,
  x = GENDER,
  y = MATHS,
  type = "robust",
  title = "Robust: Yuen's trimmed means t-test",
  messages = FALSE,
  palette = "Set2",
  package = "RColorBrewer"
) + theme_bw()
```

#### (iv) Bayesian t-test
$H_0$: $\mu_{Female} = \mu_{Male}$

$H_1$: $\mu_{Female} \neq \mu_{Male}$

$$B_{10} = \frac{P(Data | H_1)}{P(Data | H_0)}$$

```{r}
# Bayesian
ggbetweenstats(
  data = exam,
  x = GENDER,
  y = MATHS,
  type = "bayes",
  title = "Bayesian: Bayesian t-test",
  messages = FALSE,
  palette = "Set2",
  package = "RColorBrewer"
) + theme_bw()
```
:::

### (c) 1-way ANOVA

$H_0$: $\mu_{Chinese} = \mu_{Indian} = \mu_{Malay} = \mu_{Others}$ (Mean English scores of all races are equal)

$H_1$: \text{At least one } $\mu$ \text{ is different} (Not all mean English scores are equal)

```{r}
ggbetweenstats(
  data = exam,
  x = RACE, 
  y = ENGLISH,
  type = "p",
  mean.ci=TRUE,
  pairwise.comparisons = TRUE, 
  pairwise.display = "s",  # 'ns': shows only non-sig, 's': shows only sig, 'all': both 
  p.adjust.method = "fdr",
  messages = FALSE
) +
  theme_bw()  
```
::::


### 3.2 Correlation

* In the code chunk below, [ggscatterstats()](https://www.indrapatil.com/ggstatsplot/reference/ggscatterstats.html) is used to build a visual for Significant Test of Correlation between Maths scores and English scores.
* `ggscatterstats()` doesn't have a direct `color` argument for the points. Use point.args instead to pass colour settings to the underlying `geom_point()`


$H_0$: $\rho = 0$ (There is no correlation between Math and English scores)

$H_1$: $\rho \neq 0$ (There is a significant correlation between Math and English scores)

```{r}
ggscatterstats(
  data = exam,
  x = MATHS,
  y = ENGLISH,
  type = "nonparametric",
  marginal = FALSE,
  point.args = list(color = "rosybrown",
                    size = 2,
                    alpha = 0.6)
) + theme_bw()
```

### 3.3 Association

$H_0$: MATHS_bins and GENDER are independent

$H_1$: MATHS_bins and GENDER are not independent

* we need to mutate the data into bins first using `cut`.
* From the results, since p-value > 0.05 thus we have not enough statistical evidence to reject the null hypothesis that there is not association between the math_bin and gender variables.
```{r}

exam1 <- exam %>% 
  mutate(MATHS_bins = 
           cut(MATHS, 
               breaks = c(0,60,75,85,100))
)

ggbarstats(exam1,
            x=MATHS_bins,
            y=GENDER) +
  
  theme_bw() 

```



## 4 Visualising Models

* Import libraries and data (`ToyotaCorolla.csv`)
```{r}
pacman::p_load(performance, parameters, see)
car_resale <- read_csv("data/ToyotaCorolla.csv")
```

### 4.1 Multiple Regression Model using lm()

* The code chunk below is used to calibrate a multiple linear regression model by using lm() of Base Stats of R.

```{r}
## transform features
car_resale <- car_resale %>%
  mutate(
    FuelType = as.factor(FuelType),
    MetColor = as.factor(MetColor),
    Automatic = as.factor(Automatic),
    Doors = as.factor(Doors)
  )

## build model
model <- lm(Price ~ Age + 
              KM + 
              FuelType +
              HP +
              MetColor +
              Automatic +
              CC +
              Doors +
              Weight,
            data = car_resale)
```

### 4.2 Model Diagnostics Visualisations

:::: {.panel-tabset}

### (a) multicollinearity

* In the code chunk, [check_collinearity()](https://easystats.github.io/performance/reference/check_collinearity.html) of [performance](https://easystats.github.io/performance/) package.
```{r}
check_collinearity(model)
```

* render into chart
```{r}
check_c <- check_collinearity(model)
plot(check_c) + 
  theme_bw() 
```

### (b) normality

```{r}
plot(check_normality(model)) + theme_bw()
```

### (c) heteroscedasticity
In the code chunk, [check_heteroscedasticity()](https://easystats.github.io/performance/reference/check_heteroscedasticity.html) of [performance](https://easystats.github.io/performance/) package.

Heteroscedasticity refers to a situation where the variance of the **errors (or residuals)** in the linear regression model is not constant across different levels of the predictor variable(s).

```{r}
plot(check_heteroscedasticity(model)) + theme_bw()
```

### (d) complete check

* We can also perform the complete check by using [check_model()](https://easystats.github.io/performance/reference/check_model.html).
```{r}
#| fig-width: 14
#| fig-height: 12
check_model(model)
```

:::: 


### 4.3 Regression Parameters Visualisations
:::: {.panel-tabset}

### (a) see() method

In the code below, `plot()` of **see** package and `parameters()` of **parameters** package are used to visualise the parameters of a regression model.

```{r}
plot(parameters(model)) + theme_bw()
```

### (b) ggcoefstats() method

In the code below, [ggcoefstats()](https://www.indrapatil.com/ggstatsplot/reference/ggcoefstats.html) of ggstatsplot package to visualise the parameters of a regression model.

* `only.significant = TRUE` shows labels only for variables with p-value < 0.05, reducing clutter significantly!

*  Alternatively, we can replace `only.significant = TRUE`with `stats.labels = FALSE`. This removes the cluttered statistical annotation boxes and just shows the coefficient dots and confidence interval lines cleanly.
* Removing non-significant variables can improve the colour contrast of the labels that would otherwise be scaled according to the significance level.

```{r}
model2 <- lm(Price ~ Age + 
               KM + 
               FuelType +
               HP +
               CC +
               Weight,
             data = car_resale)

ggcoefstats(model2, 
            output = "plot",
            stats.label.args = list(size = 3)) +   
  theme_bw() +
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 12))
```

::::


## 5 References

* Kam, T.S. (2023). [Visual Statistical Analysis](https://r4va.netlify.app/chap10). 