---
title: "Take-Home Exercise 2: Vast Challenge 2026 - MC1"
author: "Andy Tai"
date: "05/13/2026"
date-modified: last-modified
format: 
  html:
    theme: lumen
    toc: true
    toc-depth: 3
    number-sections: false
    html-math-method: katex
    css: [styles.css, styles_mc1.css]
editor: source
execute: 
  eval: true
  echo: true
  warning: false
  message: false
  freeze: true
---

## 1 Overview
This analysis investigates the communications and decision-making processes surrounding an inappropriate information release at TenantThread, a fictional PropTech company in 2046. The dataset comprises 23 rounds of agent communications from May 17 to June 5, 2046 — spanning normal operations, an undeclared merger embargo, and a full crisis day.

Seven AI agents communicate across six channels, from open team broadcasts to private side huddles and anonymous posts. Each message may include an internal_state field capturing private agent reasoning — providing direct evidence of the gap between declared and actual behaviour.

Three questions guide the analysis: 

* what caused the release (Q1)? 
* how did agent behaviour deviate from baseline (Q2)? and 
* were there leading indicators that went unaddressed (Q3)?

## 2 Importing libraries, understanding and processing data

:::: {.panel-tabset}

### (a) Install libraries

We shall import tidyverse and other useful R libraries to aid our analysis.

```{r}
#| echo: true
#| code-fold: show 
#| message: false
#| results: hide

pacman::p_load(tidyverse, patchwork, ggrepel, ggiraph, DT, 
  gifski, plotly, gganimate, ggpp, ggtext, ggdist, ggridges, colorspace, geomtextpath, nord, nortest, seriation, dendextend, heatmaply, ggstatsplot, jsonlite, janitor, listviewer, gt, lubridate, scales, tidytext, proxy, cluster, wordcloud2, htmlwidgets, ggwordcloud, stringr, DiagrammeR, tidygraph, ggraph, igraph, kableExtra, widyr, text2vec) 
```

### (b) Import & understand data

As json file structure - comprising of key-value pairs - can be complex with many levels, we write a function that determine the json depth.

#### (i) Finding how deep is our data

Json data appears to have 5 levels of key-value pairs.
```{r}
raw <- fromJSON("data/vast2026/MC1_final_00.json")

## find how many levels
## -------------------------
# Function to calculate max depth of a nested list
max_depth <- function(x) {
  if (!is.list(x) || length(x) == 0) return(0)
  1 + max(sapply(x, max_depth))
}

cat('max. depth: ', max_depth(raw))   # 5 <-- tells you exactly how deep to set max.level
```


#### (ii) Exploring the json file structure using jsonedit from listviewer

Next, we use `jsonedit` from `listviewer` library to explore the json data using an interactive JSON viewer. We can simply click on the 'little triangle' to expand the various key(s) and explore their corresponding value(s).
```{r}
#| echo: true
jsonedit(raw)
```
#### (iii) key components of the data
The following points are gathered after exploring the data:

* the data the captured internal communications from 2 weeks leading to the breach comprises 4 main components, each with 23 unique items.

* 1) **`hour`** - contains the yyyy-mm-dd T hh:mm:ss. The year 2046 appears to be a fictional future scenario.

* 2) **`environment_context`** - captures environmental sensing data, and they comprises the following sub-components:
  * `event_narrative` and `event_headline` - appears to be key events captured during the period.
  * `market_snapshot` - tracks movement, changes, and sentiments of company stock price.
  * `media_events` - tracks key events and annoucements in the industry.
  * `social_state`- refers to scans of social media feed linked to the company.
  * `news` - appear to be news directly related to the company.
  * Others: `external_actor_actions`, `social_manager_alerts`, `agents_unavailable`, `critical_deadlines`.
  
* 3) **`communications`** - captures the message id, specific agents, communication channels and internal state. The most interesting ones are:
  * `internal_state` / `deliberating` -  as it captures the agents' thinking process.
  * `content` - appears to be minutes of meeting captured in comms' huddle.
  
* 4) **`participants`** - captures specific agents data, such as their id and roles.

#### (iv) Deep Search for all players
Not all players are apparent in the json file as a few of them only appear later in the time horizon. Hence, we write a function that searches through all unique players in the json file.
```{r}
raw <- fromJSON("data/vast2026/MC1_final_00.json", simplifyVector = FALSE)

messages_tbl <- map_dfr(raw$rounds, function(r) {
  round_hour <- ymd_hms(r$hour)
  map_dfr(r$communications, function(m) {
    ist <- m$internal_state %||% list()
    tibble(
      round_hour   = round_hour,
      timestamp    = ymd_hms(m$timestamp),
      agent_id     = m$agent_id     %||% NA_character_,
      agent_label  = m$agent_label  %||% NA_character_,
      channel      = m$channel      %||% NA_character_,
      message_type = m$message_type %||% NA_character_,
      recipients   = list(unlist(m$recipients %||% list("ALL"))),
      content      = m$content      %||% NA_character_,
      deliberating = ist$deliberating %||% NA_character_,
      reacting     = ist$reacting     %||% NA_character_,
      rationalizing= ist$rationalizing %||% NA_character_
    )
  })
})

messages_tbl %>%
  distinct(agent_id) %>%
  arrange(agent_id)
```

In summary, the players are:

* `intern_agent` - general intern.
* `judge_agent` - Compliance Officer, in charge of evaluating risks, mediates conflicts, and provides compliance guidance. Incidentally, this was mislabeled as `judge_eval_agent` in the data description document.
* `legal_agent` - General Legal counsel.
* `pr_agent` - Head of Communications and Public Relationships, in charge of PR
* `pr_intern_agent` - PR team intern with access to the official TenantThread Flex account.
* `quality_agent` - VP of Platform Trust & Safety.
* `social_media_agent` - Social-Manager in charge of social media messaging

#### (v) Important dates
There are three important dates - start date, crisis date (where embargo ends), and end date - and we shall write a function to help us search through the data for these exact dates and time. The embargo date is also an important event, and it shall be analysed in greater detail in the section 3.

```{r}
# Start date: first round in the dataset
start_date <- ymd_hms(raw$rounds[[1]]$hour)

## Embargo was first mentioned in official channel (comms_huddle)
## -------------------------------------------------------------
## Find formal embargo declaration — announced in comms_huddle
# map_dfr(raw$rounds, function(r) {
#   map_dfr(r$communications, function(m) {
#     content <- str_to_lower(m$content %||% "")
#     if (str_detect(content, "embargo") &
#         m$channel %in% c("comms_huddle", "official_post")) {
#       tibble(
#         hour    = ymd_hms(r$hour),
#         agent   = m$agent_id %||% NA_character_,
#         channel = m$channel  %||% NA_character_,
#         content = str_trunc(m$content %||% "", 300)
#       )
#     }
#   })
# }) %>%
#   arrange(hour)
# 
# embargo_date <- map_dfr(raw$rounds, function(r) {
#   round_hour <- ymd_hms(r$hour)
#   has_embargo <- any(map_lgl(r$communications, function(m) {
#     content <- str_to_lower(m$content %||% "")
#     # only count public announcement, not private deliberation
#     str_detect(content, "embargo") &
#       m$channel %in% c("comms_huddle", "official_post")
#   }))
#   tibble(hour = round_hour, has_embargo = has_embargo)
# }) %>%
#   filter(has_embargo) %>%
#   slice_min(hour, n = 1) %>%
#   pull(hour)

## Embargo date: first round where "embargo" appears in content or deliberating
## ------------------------------------------------------------------------------
# embargo_date <- map_dfr(raw$rounds, function(r) {
#   round_hour <- ymd_hms(r$hour)
#   has_embargo <- any(map_lgl(r$communications, function(m) {
#     content <- str_to_lower(m$content %||% "")
#     delib   <- str_to_lower(
#       (m$internal_state %||% list())$deliberating %||% ""
#     )
#     str_detect(content, "embargo") | str_detect(delib, "embargo")
#   }))
#   tibble(hour = round_hour, has_embargo = has_embargo)
# }) %>%
#   filter(has_embargo) %>%
#   slice_min(hour, n = 1) %>%
#   pull(hour)

# Crisis date: first round where multiple hourly rounds occur on same day
# (Crisis Day has hourly rounds vs daily rounds pre-crisis)
crisis_date <- map_dfr(raw$rounds, function(r) {
  tibble(hour = ymd_hms(r$hour), date = as.Date(ymd_hms(r$hour)))
}) %>%
  count(date) %>%
  filter(n > 1) %>%
  slice_min(date, n = 1) %>%
  mutate(hour = ymd_hms(paste(date, "09:00:00"))) %>%
  pull(hour)


# End date: last round in the dataset
end_date <- ymd_hms(raw$rounds[[length(raw$rounds)]]$hour)

# ── Print to verify ──────────────────────────────────────────────────────────
cat("Start date:  ", format(start_date,   "%d %b %Y %H:%M"), "\n")
#cat("Embargo date:", format(embargo_date, "%d %b %Y %H:%M"), "\n")
cat("Crisis date: ", format(crisis_date,  "%d %b %Y %H:%M"), "\n")
cat("End date:    ", format(end_date,     "%d %b %Y %H:%M"), "\n")
```

### (c) Data Wrangling

The data file is in json format, and analysing that is tricky given that many key pieces of information that are related to one another are stored in different layers. Hence, we first identify the key pieces and lay them out in a flat table to facilitate analysis.

The following code checks for potential problems - namely, missing values and duplicative data - and then proceed to filter the data based on 3 conditions:

* `sentiment` contains either "negative" or "critical" tags.
* `percent_change` has a -0.5 effect on stock price.
* `narrative` contains key words that contain negative signals.

Essentially, we want to extract the key events - especially breaches and leak signals - that impact the stock prices of TenantThread negatively. We note that there are altogether five key events that satisfy the above-mentioned criteria.

```{r}
# ============================================================
# LOAD JSON
# ============================================================

`%||%` <- function(x, y) if (is.null(x)) y else x

raw <- fromJSON(
  "data/vast2026/MC1_final_00.json",
  simplifyVector = FALSE
)

# ============================================================
# 1. EXTRACT MARKET + EVENT NARRATIVE DATA
# ============================================================

stock_tbl <- map_dfr(raw$rounds, function(r) {
  
  ms <- r$environment_context$market_snapshot
  
  tibble(
    hour = ymd_hms(r$hour), ## convert to datetime
    price = parse_number(ms$stock_price %||% NA_character_),
    percent_change = parse_number(ms$percent_change %||% NA_character_),
    sentiment = str_to_lower(ms$sentiment %||% NA_character_),
    headline = r$environment_context$event_headline %||% NA_character_,
    narrative = r$environment_context$event_narrative %||% NA_character_
  )
  
}) %>%
  mutate(t = row_number())

# ============================================================
# 2. DATA QUALITY CHECKS
# ============================================================

missing_check <- stock_tbl %>%
  summarise(
    rows = n(),
    missing_hour = sum(is.na(hour)),
    missing_price = sum(is.na(price)),
    missing_percent_change = sum(is.na(percent_change)),
    missing_sentiment = sum(is.na(sentiment)),
    missing_narrative = sum(is.na(narrative))
  )

duplicate_check <- stock_tbl %>%
  count(hour) %>%
  filter(n > 1)

missing_check
duplicate_check


## ==========================================
## 3. CREATE LEAKAGE SIGNAL PATTERNS DETECTOR
## ==========================================

## sentiment %in% c("negative", "critical") ## <-- CONDITION 1
## percent_change < -0.5  ## <- CONDITION 2 (HAS IMPACT ON STOCK PRICE)

## CONDITION 3 - CONTAINS PARTICULAR THEMATIC KEYWORDS
theme_dict <- tribble(
  ~theme, ~pattern,
  "Disclosure Risk", "leak|leakage|embargo",
  "Anonymous Activity", "anonymous",
  "Media Scrutiny", "journal|reporter|investigation",
  "Privacy Exposure", "re-identification",
  "Financial Distress", "covenant breach|distressed|sale",
  "Merger / M&A", "merge|merger|acquisition"
)

## =================================
## 4. FILTERING BASED ON CONDITIONS
## =================================
risk_theme_tbl <- stock_tbl %>%
  ## handles missing values before mutate
    filter(
    !is.na(price),
    !is.na(percent_change),
    !is.na(sentiment),
    !is.na(narrative)
  ) %>%
  mutate(
    narrative_lower = str_to_lower(narrative),
    bad_sentiment = sentiment %in% c("negative", "critical"),
    large_drop = percent_change < -0.5
  ) %>%
  tidyr::crossing(theme_dict) %>%
  mutate(
    theme_match = str_detect(narrative_lower, pattern)
  ) %>%
  filter(
    bad_sentiment,
    large_drop,
    theme_match
  ) %>%
  group_by(hour, price, percent_change, sentiment, headline, narrative) %>%
  summarise(
    themes = paste(unique(theme), collapse = "; "),
    .groups = "drop"
  ) %>%
  arrange(hour)

risk_theme_tbl
```

::::

## 3 Visualise sequence of key events and relationships that led up to inappropriate release

### What this section is about

Here we look at how a combination of organisational, communication and governance failures led to the inappropriate disclosure of TenantThread’s merger plans. The individual tabs look at the details of the implementation of the embargo, key precursor events, the sequence of events leading to the breach, the effect on the stock price, and the communication patterns between the agents in the days leading up to Crisis Day.

### Section summary

Evidence suggests that merger disclosure was not the work of a single rogue actor. Instead, it was the culmination of a series of failures of governance and communication that preceded the breach. There was never a formal announcement of the embargo, critical information was conveyed informally, and many warning signs were missed. The weaknesses meant that sensitive merger information could flow through the organisation and was subsequently made public, leading to a loss of investor confidence.

:::: {.panel-tabset}

### (a) implementation of embargo

The interactive data table below is gathered based on a deep keyword search through the json data for all mentions of 'embargo'. A few points stand out:

* Besides the standard communication channels: 
  * There are 2 types of public posts: "personal post", "anonymous post"
  * Besides the 3 types of internal communication channels, which are namely: "comms huddle", "one-on-one chat", and "side-huddle", there is a 1 type of hidden communication channel that is not captured in the data file. We call this "Private / Direct Messaging". This can only be inferred from "internal_state/deliberating", and it was how CEO Ajay communicated the need for "embargo" to the Legal-Agent on 23-05-2046 at 09:00. We refer this date as the "official" date of the embargo.
  
* While the the need and intent of the embargo was communicated to the Legal-Agent, he never did officially declare this company-wide. Instead, he used one-on-one channel and side-huddle to communicate it to the senior management staff. The Legal-Agent did so presumably for two reasons: 1) he communicated to essential senior staff on a need to know basis, and 2) officially declaring an embargo may invite unwanted speculations.

* The Legal-Agent, however, did took control of the communications flow in the company, as evidenced in the network graph, as communications appear to flow from and to the "Legal" node.

```{=html}
<style>
  #embargo-table tbody tr:hover {
    background-color: #d6eaf8 !important;
    cursor: default;
  }
  #embargo-table thead tr th {
    background-color: #2c3e50;
    color: white;
    padding: 8px;
  }
  #embargo-table table { 
    border-collapse: collapse; 
    font-size: 10px; 
  }
  #embargo-table td { 
    border-bottom: 1px solid #ddd; 
    padding: 6px; 
    vertical-align: top;
  }
  .col-nowrap { white-space: nowrap; overflow: hidden; text-overflow: ellipsis; }
  .col-wrap   { white-space: normal; word-wrap: break-word; }
</style>
```

::: {#embargo-table}
```{r embargo-table-search}
#| echo: false

embargo_mentions <- map_dfr(raw$rounds, function(r) {
  map_dfr(r$communications, function(m) {
    ist     <- m$internal_state %||% list()
    content <- m$content        %||% ""
    delib   <- ist$deliberating %||% ""
    react   <- ist$reacting     %||% ""
    ratio   <- ist$rationalizing %||% ""
    all_text <- paste(content, delib, react, ratio)
    if (!str_detect(str_to_lower(all_text), "embargo")) return(NULL)
    tibble(
      timestamp = ymd_hms(m$timestamp),
      from      = m$agent_label %||% NA_character_,
      to        = case_when(
        length(unlist(m$recipients %||% list())) > 0 &
          !all(unlist(m$recipients %||% list()) == "ALL") ~
          paste(unlist(m$recipients), collapse = ", "),
        m$channel == "comms_huddle"    ~ "All team",
        m$channel == "one_on_one_chat" ~ "(private)",
        m$channel == "side_huddle"     ~ "(small group)",
        m$channel == "personal_post"   ~ "(public external)",
        m$channel == "anonymous_post"  ~ "(public anonymous)",
        TRUE ~ "ALL"
      ),
      channel         = m$channel %||% NA_character_,
      found_in        = case_when(
        str_detect(str_to_lower(content), "embargo") &
          str_detect(str_to_lower(delib), "embargo") ~ "content + deliberating",
        str_detect(str_to_lower(content), "embargo") ~ "content",
        str_detect(str_to_lower(delib),   "embargo") ~ "deliberating",
        str_detect(str_to_lower(react),   "embargo") ~ "reacting",
        str_detect(str_to_lower(ratio),   "embargo") ~ "rationalizing",
        TRUE ~ "other"
      ),
      content_excerpt = content,
      delib_excerpt   = delib
    )
  })
}) %>%
  arrange(timestamp) %>%
  select(timestamp, from, to, channel,
         found_in, content_excerpt, delib_excerpt)

datatable(
  embargo_mentions,
  rownames = FALSE,
  filter   = "top",
  options  = list(
    pageLength   = 5,
    lengthMenu   = c(5, 10, 20),
    scrollX      = TRUE,
    scrollY      = "400px",
    autoWidth    = FALSE,
    initComplete = JS("
      function(settings, json) {
        var api = this.api();
        api.columns(0).nodes().to$().css('width', '110px');
        api.columns(1).nodes().to$().css('width', '90px');
        api.columns(2).nodes().to$().css('width', '90px');
        api.columns(3).nodes().to$().css('width', '90px');
        api.columns(4).nodes().to$().css('width', '80px');
        api.columns(5).nodes().to$().css('width', '300px');
        api.columns(6).nodes().to$().css('min-width', '400px');
        api.columns.adjust().draw();
      }
    "),
    columnDefs = list(
      list(targets = 0, width = "110px"),
      list(targets = 1, width = "90px"),
      list(targets = 2, width = "90px"),
      list(targets = 3, width = "90px"),
      list(targets = 4, width = "80px"),
      list(targets = 5, width = "300px"),
      list(targets = 6, width = "400px")
    )
  ),
  caption = "All embargo mentions across all rounds and fields"
)
```
:::

```{r embargo-date-assign}
# Conclusion from Section 3(a) analysis above
# Earliest communication: May 23 via side_huddle by Legal
# No formal comms_huddle announcement was ever made
embargo_date <- ymd_hms("2046-05-23T09:00:00")
cat("Embargo date set to:", format(embargo_date, "%d %b %Y %H:%M"), "\n")
```

### (b) investigate precursors

Aside from the table in tab (a) that zoom in on all mention of 'embargo' in the dataset, we also need a more exhaustive search for other strategic keywords so that we can build up a proper chain of key events leading up to the leakage of information. Hence, another interactive data table is created to curate these relevant messages.

```{r}
# Evidence extractor — search any keyword across all message fields
search_messages <- function(pattern, label = "") {
  map_dfr(raw$rounds, function(r) {
    map_dfr(r$communications, function(m) {
      ist     <- m$internal_state %||% list()
      content <- m$content        %||% ""
      delib   <- ist$deliberating %||% ""
      react   <- ist$reacting     %||% ""
      all_text <- str_to_lower(paste(content, delib, react))
      
      if (!str_detect(all_text, pattern)) return(NULL)
      
      tibble(
        signal          = label,
        timestamp       = ymd_hms(m$timestamp),
        from            = m$agent_label %||% NA_character_,
        channel         = m$channel     %||% NA_character_,
        found_in        = case_when(
          str_detect(str_to_lower(content), pattern) ~ "content",
          str_detect(str_to_lower(delib),   pattern) ~ "deliberating",
          TRUE ~ "reacting"
        ),
        content_excerpt = content,
        delib_excerpt   = delib
      )
    })
  }) %>% arrange(timestamp)
}

## filter based on bespoke keywords
evidence_tbl <- bind_rows(
  search_messages("strategic development|career.defining",
                  "Strategic signal"),
  search_messages("shadow channel",
                  "Shadow channel"),
  search_messages("whatever i share|interns cannot be briefed",
                  "CEO embargo instruction"),
  search_messages("social hold|no.*review|monitor.*not yet",
                  "No enforcement"),
  search_messages("elena|elenamarquez",
                  "Elena post"),
  search_messages("civicloom.*like|like.*civicloom",
                  "CivicLoom exposure"),
  search_messages("compliance monitor.*join|introducing.*monitor",
                  "Monitor assigned")
)
```

#### Interactive Data Table - to search through all strategic messages

```{r}
datatable(
  evidence_tbl,
  rownames = FALSE,
  filter   = "top",
  options  = list(
    pageLength   = 5,
    lengthMenu   = c(5, 10, 20),
    scrollX      = TRUE,
    autoWidth    = FALSE,
    initComplete = JS("
      function(settings, json) {
        var api = this.api();
        // set font size on entire table
        $(api.table().node()).css('font-size', '10px');
        // force column widths
        api.columns(0).nodes().to$().css('min-width', '80px').css('max-width', '80px');
        api.columns(1).nodes().to$().css('min-width', '90px').css('max-width', '90px');
        api.columns(2).nodes().to$().css('min-width', '80px').css('max-width', '80px');
        api.columns(3).nodes().to$().css('min-width', '80px').css('max-width', '80px');
        api.columns(4).nodes().to$().css('min-width', '80px').css('max-width', '80px');
        api.columns(5).nodes().to$().css('min-width', '250px');
        api.columns(6).nodes().to$().css('min-width', '400px');
        api.columns.adjust().draw();
      }
    "),
    columnDefs = list(
      list(targets = 0, width = "60px"),
      list(targets = 1, width = "60px"),
      list(targets = 2, width = "60px"),
      list(targets = 3, width = "60px"),
      list(targets = 4, width = "60px"),
      list(targets = 5, width = "350px"),
      list(targets = 6, width = "500px")
    )
  ),
  caption = "Evidence table — raw data supporting each precursor signal"
) %>%
  formatStyle(
    columns    = 0:6,
    fontSize   = "10px",
    whiteSpace = "normal"
  )
```

### (c) visualise key events
This flow diagram illustrates key events, including the precursors, triggering events, feedback mechanisms, and how they collectively set the domino effect in motion. The information for this was obtained from viewing the interactive data tables in part (a) and (b).

The key highlights are:

* May 17 — Structural gaps existed from day one: no posting review workflow and merger speculation already circulating with no controls triggered.
* May 22 — Sensitive merger discussions moved to an unaudited shadow channel, bypassing normal team oversight.
* May 23 —  Embargo obligation created informally. CEO Ajay instructed Legal via a private DM outside the logged system. Legal understood this as an embargo instruction but never formally communicated it to the full team, and no technical posting restrictions were implemented.
* May 29 —  Social Media agent breached the embargo by posting publicly and tagging external PropTech figure @ElenaMarquez with caption "big things coming". CivicLoom saw and liked the post within minutes before Legal had it deleted 14 minutes later.
* May 30 — Judge assigned reactively, after the breach had already occurred and the signal had reached the merger counterparty.


```{r precursor-extract}
#| include: false

# ── Step 1: flatten all messages + deliberating into one searchable table ───
all_text_tbl <- map_dfr(raw$rounds, function(r) {
  map_dfr(r$communications, function(m) {
    ist <- m$internal_state %||% list()
    tibble(
      round_hour  = ymd_hms(r$hour),
      timestamp   = ymd_hms(m$timestamp),
      agent_id    = m$agent_id    %||% NA_character_,
      agent_label = m$agent_label %||% NA_character_,
      channel     = m$channel     %||% NA_character_,
      content     = m$content     %||% NA_character_,
      deliberating= ist$deliberating %||% NA_character_
    )
  })
}) %>%
  mutate(
    combined = str_to_lower(
      paste(coalesce(content, ""), coalesce(deliberating, ""))
    )
  )

# ── Step 2: precursor signal dictionary ────────────────────────────────────
# Each signal has a label, search pattern, decision description, and
# what system gap it represents
precursor_dict <- list(

  "Strategic signal" = list(
    pattern       = "strategic development|career.defining|merger",
    decision      = "Merger hinted — no embargo yet",
    gap           = "Information asymmetry created",
    node_colour   = "#E8E8E8",
    border_colour = "#999999"
  ),

  "No enforcement" = list(
    pattern       = "social hold|no.*review|monitor.*not yet|no pre",
    decision      = "No pre-posting review workflow",
    gap           = "Process gap",
    node_colour   = "#FFE0B2",
    border_colour = "#E65100"
  ),

  "Shadow channel" = list(
    pattern       = "shadow channel",
    decision      = "Sensitive comms moved off-record",
    gap           = "No audit trail",
    node_colour   = "#FFF9C4",
    border_colour = "#F9A825"
  ),

  "Embargo declared" = list(
    pattern       = "whatever i share|interns cannot be briefed",
    decision      = "CEO instructs Legal via private DM",
    gap           = "No technical lock",
    node_colour   = "#FFE0B2",
    border_colour = "#E65100"
  ),

"Social Media breach" = list(
    pattern       = "elena|elenamarquez",
    decision      = "Social Media posts publicly tagging @ElenaMarquez",
    gap           = "14 min exposure",
    node_colour   = "#EF9A9A",
    border_colour = "#B71C1C"
  ),
  "CivicLoom exposure" = list(
    pattern       = "civicloom.*like|like.*civicloom",
    decision      = "CivicLoom sees the signal",
    gap           = "External leak confirmed",
    node_colour   = "#EF9A9A",
    border_colour = "#B71C1C"
  ),

  "Monitor assigned" = list(
    pattern       = "compliance monitor.*join|introducing.*monitor",
    decision      = "Judge assigned — reactively",
    gap           = "Too late",
    node_colour   = "#BBDEFB",
    border_colour = "#1565C0"
  )

)

# Convert named list to tibble BEFORE the crossing step
precursor_tbl <- imap_dfr(precursor_dict, function(val, name) {
  as_tibble(val) %>% mutate(signal = name)
})

# ── Step 3: detect which signals appear in which rounds, with top agent ─────
precursor_rounds_tbl <- all_text_tbl %>%
  tidyr::crossing(precursor_tbl %>% select(signal, pattern, decision,
                                   gap, node_colour, border_colour)) %>%
  filter(str_detect(combined, pattern)) %>%
  group_by(round_hour, signal, decision, gap, node_colour, border_colour) %>%
  summarise(
    # pick the agent with most mentions of the pattern in this round
    top_agent = {
      ag <- agent_label[str_detect(combined, pattern)]
      names(sort(table(ag), decreasing = TRUE))[1]
    },
    n_mentions = sum(str_detect(combined, pattern)),
    .groups = "drop"
  ) %>%
  # keep only the FIRST round each signal appears (the origin, not the echo)
  group_by(signal) %>%
  slice_min(round_hour, n = 1) %>%
  ungroup() %>%
  arrange(round_hour) %>%
  # join stock price at that round
  left_join(
    stock_tbl %>% select(hour, price, sentiment, percent_change),
    by = c("round_hour" = "hour")
  ) %>%
  mutate(
    node_id   = paste0("P", row_number()),
    date_lbl  = format(round_hour, "%b %d %H:%M"),
    # flag if this round is ALSO in consequence layer
    is_consequence = round_hour %in% risk_theme_tbl$hour
  )
```

---

#### Flow Diagram

*Each node is the first round a precursor signal appeared, detected from
communications content and agent deliberating fields.
Orange diamonds = system gaps. Red = the leak event.*
```{r precursor-flow}
#| fig.height: 6
#| fig.width: 14

make_precursor_dot <- function(tbl) {
  
  n        <- nrow(tbl)
  row1_ids <- tbl$node_id[1:4]
  row2_ids <- rev(c(tbl$node_id[5:n], "BRIDGE"))
  
  # ── x positions: wider spacing to avoid diamond overlap ──────────────────
  n_row1    <- ceiling(n / 2)
  n_row2    <- n - n_row1 + 1      # +1 for BRIDGE
  row1_ids  <- tbl$node_id[1:n_row1]
  row2_ids  <- rev(c(tbl$node_id[(n_row1 + 1):n], "BRIDGE"))

  x_pos_row1 <- seq(0, by = 5.5, length.out = n_row1)
  x_pos_row2 <- seq(0, by = 5.5, length.out = length(row2_ids))
  y_row1 <- 4
  y_row2 <- 0
  
  pos_row1 <- map2_chr(row1_ids, x_pos_row1, function(id, x) {
    glue::glue('  {id} [pos = "{x},{y_row1}!"]')
  }) %>% paste(collapse = "\n")
  
  pos_row2 <- map2_chr(row2_ids, x_pos_row2, function(id, x) {
    glue::glue('  {id} [pos = "{x},{y_row2}!"]')
  }) %>% paste(collapse = "\n")
  
  # ── node labels and styles ────────────────────────────────────────────────
  node_styles <- tbl %>%
    pmap_chr(function(node_id, date_lbl, signal, decision, gap,
                      top_agent, node_colour, border_colour,
                      is_consequence, ...) {
      shape <- if (str_detect(gap, "gap|lock|trail")) "diamond" else "rectangle"
      extra <- if (is_consequence) ", penwidth = 2.5" else ""
      lbl   <- paste0(
        date_lbl, "\\n",
        str_replace_all(signal, "'", " "), "\\n",
        str_wrap(decision, 18) %>% str_replace_all("\n", "\\\\n"), "\\n",
        "[", str_replace_all(top_agent %||% "", "_agent", ""), "]"
      )
      # diamonds need smaller width than rectangles
      w <- if (shape == "diamond") "2.0" else "2.8"
      h <- if (shape == "diamond") "1.2" else "1.5"
      glue::glue(
        "  {node_id} [label = '{lbl}', shape = {shape},",
        " fillcolor = '{node_colour}', color = '{border_colour}',",
        " fontsize = 13, width = {w}, height = {h}{extra}]"
      )
    }) %>%
    paste(collapse = "\n")
  
  # ── bridge node ───────────────────────────────────────────────────────────
  bridge_style <- glue::glue(
    "  BRIDGE [label = 'Consequence layer\\n(3-condition filter)',",
    " shape = rectangle, fillcolor = '#B71C1C', fontcolor = white,",
    " color = '#B71C1C', penwidth = 3, fontsize = 12,",
    " width = 3.2, height = 1.8]"
  )
  
  # ── sequential edges ──────────────────────────────────────────────────────
  edges <- tbl %>%
    mutate(next_id = lead(node_id)) %>%
    filter(!is.na(next_id)) %>%
    pmap_chr(function(node_id, next_id, gap, ...) {
      style <- if (str_detect(gap, "gap|lock")) "dashed" else "solid"
      # shorten edge labels to avoid overlap
      short_gap <- str_trunc(gap, 20)
      glue::glue(
        "  {node_id} -> {next_id} [label = '{short_gap}',",
        " style = {style}, fontsize = 13]"
      )
    }) %>%
    paste(collapse = "\n")
  
  # ── row connector: drop from end of row1 to start of row2 ────────────────
  row2_start <- tail(row2_ids, 1)   # rightmost node of row2 (reversed)
  row_connector <- glue::glue(
    "  {tail(row1_ids, 1)} -> {row2_start}",
    " [fontsize = 13]"
  )
  
  # ── bridge edge ───────────────────────────────────────────────────────────
  bridge_edge <- glue::glue(
    "  {tail(tbl$node_id, 1)} -> BRIDGE",
    " [label = 'triggers', color = '#B71C1C',",
    " penwidth = 2, fontsize = 13]"
  )
  
  glue::glue("
digraph precursor_chain {{

  graph [layout = neato, bgcolor = '#FAFAFA',
         splines = polyline, overlap = scale]

  node [fontname = Helvetica, fontsize = 13, style = filled,
        shape = rectangle, margin = '0.3,0.2']

  edge [fontname = Helvetica, fontsize = 13]

{node_styles}
{bridge_style}

{pos_row1}
{pos_row2}

{edges}
{row_connector}
{bridge_edge}
}}
")
}

DiagrammeR::grViz(make_precursor_dot(precursor_rounds_tbl))
```






---

#### Two-layer summary table

The table below contains salient bite-size pointers segregating the precursor and consequence events.

```{r two-layer-table}
#| echo: false

pre_summary <- precursor_rounds_tbl %>%
  transmute(
    Layer        = "Precursor",
    Date         = format(round_hour, "%b %d"),
    Signal       = signal,
    Decision     = decision,
    `System gap` = gap,
    Actor        = str_replace_all(top_agent %||% "", "_agent", ""),
    Price        = ifelse(is.na(price), "—", dollar(price)),
    Sentiment    = str_to_title(coalesce(sentiment, "—"))
  )

con_summary <- risk_theme_tbl %>%
  transmute(
    Layer        = "Consequence",
    Date         = format(hour, "%b %d %H:%M"),
    Signal       = themes,
    Decision     = str_trunc(headline, 50),
    `System gap` = "Impact now visible",
    Actor        = "Multiple",
    Price        = dollar(price),
    Sentiment    = str_to_title(sentiment)
  )

combined_tbl <- bind_rows(pre_summary, con_summary)

datatable(
  combined_tbl,
  rownames = FALSE,
  filter   = "top",
  options  = list(
    pageLength = 5,
    lengthMenu = c(5, 10, 20),
    scrollX    = TRUE,
    autoWidth  = FALSE,
    columnDefs = list(
      list(width = "8%",  targets = 0),   # Layer
      list(width = "6%",  targets = 1),   # Date
      list(width = "12%", targets = 2),   # Signal
      list(width = "20%", targets = 3),   # Decision
      list(width = "15%", targets = 4),   # System gap
      list(width = "8%",  targets = 5),   # Actor
      list(width = "6%",  targets = 6),   # Price
      list(width = "8%",  targets = 7)    # Sentiment
    )
  ),
  caption = "Two-layer summary: precursor events feeding into consequence rounds"
) %>%
  formatStyle(
    "Layer",
    backgroundColor = styleEqual(
      c("Precursor",  "Consequence"),
      c("#FFF3E0",    "#FFEBEE")
    ),
    fontWeight = "bold"
  ) %>%
  formatStyle(
    "Sentiment",
    color = styleEqual(
      c("Neutral",  "Cautious",  "Negative",  "Critical"),
      c("#888888",  "#E65100",   "#C62828",   "#B71C1C")
    )
  )
```


### (d) impact on stock price

The stock price of TenantThread between 17 May 2046 to 05 June 2046 is plotted in the line chart below. In addition, a linear projection is superimposed on the same plot, and this is based on the organic decline patterns prior to the first leak signal being detected (which was: "2046-05-25 09:00:00 UTC"). Furthermore, we also inserted the 5 key events obtained earlier in the line plot. 

Collectively, the graph illustrates that TenantThead's stock price decline during the two-week period is not organic. The five key events played a major role in affecting shareholders' confidence.

The details of these five key events are displayed in the table A below. The salient points are:

* 25 May: Slowing growth, uncertain client renewals, and merger discussions signalled weakening business fundamentals.
* 30 May: Compliance intervention and stricter embargo controls highlighted growing disclosure and governance risks.
* 31 May: SaltWind's first article exposed data broker partnerships, triggering public scrutiny and fears of information leakage.
* 1 June: Damage-control efforts and competitor attacks reinforced concerns that the merger was TenantThread's only viable path forward.
* 4 June: SaltWind's second exposé on re-identification risks caused the sharpest stock decline and set the stage for the crisis day.

Overall: Investor confidence eroded progressively due to deteriorating fundamentals, increasing merger dependence, disclosure concerns, and mounting privacy-related media scrutiny.

#### key events on stock price

```{r}
## Fit the "organic decline" BEFORE leakage signals emerge
## ---------------------------------------------------------

first_signal <- min(risk_theme_tbl$hour)

pre_leak_tbl <- stock_tbl %>%
  filter(hour < first_signal) %>%
  mutate(
    days = as.numeric(
      difftime(
        hour,
        min(stock_tbl$hour),
        units = "days"
      )
    )
  )

organic_model <- lm(
  price ~ days,
  data = pre_leak_tbl
)

## project trend forward
stock_plot_tbl <- stock_tbl %>%
  arrange(hour) %>%
  mutate(
    days = as.numeric(
      difftime(
        hour,
        min(stock_tbl$hour),
        units = "days"
      )
    ),
    
    organic_price = predict(
      organic_model,
      newdata = data.frame(
        days = days
      )
    ),
    
    leakage_gap = price - organic_price,
    
    ribbon_low = ifelse(
      price < organic_price,
      price,
      NA
    ),
    
    ribbon_high = ifelse(
      price < organic_price,
      organic_price,
      NA
    )
  )

## prepare tooltip data
## ----------------------


risk_plotly_tbl <- risk_theme_tbl %>%
  mutate(
    date_label = format(hour, "%d %b %Y"),
    time_label = format(hour, "%H:%M"),
    price_label = dollar(price),
    change_label = sprintf("%.1f%%", percent_change),
    sentiment_label = str_to_title(sentiment),
    theme_wrapped = str_wrap(themes, width = 24),
    
    tooltip = paste0(
      "<b>Date:</b> ", date_label,
      "<br><b>Time:</b> ", time_label,
      "<br><b>Price:</b> ", price_label,
      "<br><b>Change:</b> ", change_label,
      "<br><b>Sentiment:</b> ", sentiment_label,
      "<br><b>Theme:</b> ", themes,
      "<br><b>Event:</b> ", headline
    )
  )

gap_tbl <- stock_plot_tbl %>%
  filter(price < organic_price)

# ============================================================
# PLOTLY CHART
# ============================================================

plot_ly() %>%
  
  # Actual stock price
  add_lines(
    data = stock_plot_tbl,
    x = ~hour,
    y = ~price,
    name = "Actual stock price",
    line = list(color = "black", width = 3),
    hovertemplate = paste(
      "<b>Date:</b> %{x|%d %b %Y}",
      "<br><b>Actual price:</b> $%{y:.2f}",
      "<extra></extra>"
    )
  ) %>%
  
  # Projected organic decline
  add_lines(
    data = stock_plot_tbl,
    x = ~hour,
    y = ~organic_price,
    name = "Projected organic decline",
    line = list(color = "blue", width = 3, dash = "dash"),
    hovertemplate = paste(
      "<b>Date:</b> %{x|%d %b %Y}",
      "<br><b>Organic trend:</b> $%{y:.2f}",
      "<extra></extra>"
    )
  ) %>%
  
  # Shaded deviation area between actual and organic line
  add_ribbons(
    data = stock_plot_tbl %>% filter(!is.na(ribbon_low)),
    x        = ~hour,
    ymin     = ~ribbon_low,
    ymax    = ~ribbon_high,
    name     = "Deviation from organic trend",
    fillcolor = "rgba(255, 100, 100, 0.35)",
    line      = list(color = "transparent"),
    hoverinfo = "skip"
  ) %>%
  
  # Theme bubbles
  add_markers(
    data = risk_plotly_tbl,
    x = ~hour,
    y = ~price,
    color = ~theme_wrapped,
    split = ~theme_wrapped,
    text = ~tooltip,
    hoverinfo = "text",
    marker = list(
      size = 16,
      opacity = 0.85,
      line = list(
        color = "white",
        width = 1.5
      )
    )
  ) %>%
  
  layout(
    title = list(
      text = paste0(
        "Stock Price and Potential Leakage Signals",
        "<br><sup>Theme bubbles show events satisfying sentiment, price-drop and thematic-risk conditions</sup>"
      )
    ),
    
    xaxis = list(
      title = "",
      tickformat = "%b %d"
    ),
    
    yaxis = list(
      title = "Stock Price",
      range = c(25, 40)
    ),
    
    legend = list(
      orientation = "v",
      x = 1.02,
      y = 1
    ),
    
    margin = list(
      l = 60,
      r = 260,
      t = 80,
      b = 60
    ),
    
    hoverlabel = list(
      bgcolor     = "#2c3e50",
      bordercolor = "#2c3e50",
      font        = list(color = "snow", size = 11),
      align       = "left"
    )
  )
```

#### Table A: The 5 key events that negatively impact TenantThread's stock price between 17 May 2046 to 5 June 2046

```{=html}
<style>
  #leakage-table tbody tr:hover {
    background-color: #d6eaf8 !important;
    cursor: default;
  }
  #leakage-table thead tr th {
    background-color: #2c3e50;
    color: white;
    padding: 8px;
  }
  #leakage-table table { width: 100%; border-collapse: collapse; font-size: 10px; }
  #leakage-table td    { border-bottom: 1px solid #ddd; padding: 6px; vertical-align: top; }
  #leakage-table td:nth-child(8) { width: 50%; white-space: normal; }
  #leakage-table td:nth-child(7) { width: 20%; white-space: normal; }
</style>
```

::: {#leakage-table}
```{r}
#| echo: false
risk_theme_tbl %>%
  mutate(
    Date        = format(hour, "%d %b %Y"),
    Time        = format(hour, "%H:%M"),
    Price       = dollar(price),
    `Change (%)` = sprintf("%.1f%%", percent_change),
    Sentiment   = str_to_title(sentiment),
    Theme       = themes,
    Event       = headline,
    Narrative   = narrative
  ) %>%
  select(Date, Time, Price, `Change (%)`, Sentiment, Theme, Event, Narrative) %>%
  knitr::kable(format = "html")
```
:::





### (e) post-embargo communications flow

Next, we zoom into the period between embargo and news leakage to look at the communications between various actors and the channels used. In this regard, we used a network graph to visualise the messages in `communications` between sender (based on `agent_id`) and receiver (based on `receipients`)

The following network graph is constructed based on every sender-receiver pair in the messaging data. The edges are counted for every from-to-channel pair and that becomes the edge weight driving thickness. The node size is calculated as total messages sent per agent.

Based on the network graph, we can draw the following insights:

* `Legal` appears to be directing almost everyone.
* Most communications are between `legal`, `platform trust`, and `Social Media`, and a combination of comms huddle and side huddle are used. That is to say, that some conversations are official and open, while others involve private group coordination that are not visible to the full team.
* `Intern`, `PR Intern` and `PR` received many arrows but send back few is consistent with their roles as executors as opposed to decision-makers
* The node size of `Judge` and `Intern` are relatively smaller and received thinner edges, meaning that the messages that they received are fewer, or that they are involved much later in the game.
* There appears to be a few one-on-one chat between `PR Intern` and `Intern`, `PR Intern` and `PR`, `Social Media` and `PR`, as well as between `Judge` and `Legal`, as indicated by the blue edges.

```{r network-setup}
#| include: false

## embargo_date <- ymd_hms("2046-05-23T00:00:00")

# ── Step 1: get active agents per round (needed to expand ALL) ───────────────
active_agents_per_round <- map_dfr(raw$rounds, function(r) {
  agents <- unique(map_chr(r$communications, ~.x$agent_id %||% NA_character_))
  tibble(
    hour   = ymd_hms(r$hour),
    agents = list(agents[!is.na(agents)])
  )
})

# ── Step 2: extract edges, expanding ALL for pre-embargo broadcasts ──────────
network_edges_raw <- map_dfr(raw$rounds, function(r) {
  round_hour <- ymd_hms(r$hour)
  is_pre     <- round_hour < embargo_date

  # active agents this round (for expanding ALL)
  active <- active_agents_per_round %>%
    filter(hour == round_hour) %>%
    pull(agents) %>%
    unlist()

  map_dfr(r$communications, function(m) {
    from <- m$agent_id %||% NA_character_
    recs <- unlist(m$recipients %||% list("ALL"))
    chan <- m$channel %||% NA_character_

    # expand ALL to every other active agent
    if ("ALL" %in% recs) {
      recs <- setdiff(active, from)
    }

    tibble(
      hour    = round_hour,
      from    = from,
      to      = recs,
      channel = chan,
      is_pre  = is_pre
    )
  })
}) %>%
  filter(!is.na(from), !is.na(to), from != to)

# ── Step 3: normalise recipient names to agent_id format ────────────────────
network_edges <- network_edges_raw %>%
  mutate(
    to = case_when(
      to == "legal"          ~ "legal_agent",
      to == "social_manager" ~ "social_media_agent",
      to == "platform_trust" ~ "quality_agent",
      to == "pr"             ~ "pr_agent",
      to == "pr_intern"      ~ "pr_intern_agent",
      to == "intern"         ~ "intern_agent",
      to == "judge"          ~ "judge_agent",
      TRUE                   ~ to
    ),
    period = if_else(
      is_pre,
      "Pre-Embargo (May 17-22)",
      "Post-Embargo (May 23-Jun 5)"
    )
  ) %>%
  count(period, from, to, channel, name = "weight")

# ── Step 4: agent metadata ───────────────────────────────────────────────────
agent_meta <- tribble(
  ~agent_id,            ~label,            ~colour,
  "legal_agent",        "Legal",           "#3266ad",
  "quality_agent",      "Platform Trust",  "#1D9E75",
  "social_media_agent", "Social Media",    "#D85A30",
  "pr_agent",           "PR",              "#B07EB8",
  "pr_intern_agent",    "PR Intern",       "#BA7517",
  "judge_agent",        "Judge",           "#5F5E5A",
  "intern_agent",       "Intern",          "#888780"
)

channel_colours <- c(
  "comms_huddle"    = "#1D9E75",
  "one_on_one_chat" = "#1303fc",
  "side_huddle"     = "#D85A30",
  "official_post"   = "#B07EB8",
  "personal_post"   = "#BA7517",
  "anonymous_post"  = "#5F5E5A"
)

# ── Step 5: shared plot function ─────────────────────────────────────────────
make_network_plot <- function(period_label, plot_title, plot_subtitle) {

  edges <- network_edges %>%
    filter(period == period_label)

  if (nrow(edges) == 0) return(NULL)

  nodes <- tibble(
    agent_id = unique(c(edges$from, edges$to))
  ) %>%
    left_join(agent_meta, by = "agent_id") %>%
    mutate(
      label  = coalesce(label,  agent_id),
      colour = coalesce(colour, "#888888"),
      n_sent = map_int(agent_id,
                       ~sum(edges$weight[edges$from == .x]))
    )

  g <- tbl_graph(
    nodes    = nodes,
    edges    = edges %>%
      rename(from_id = from, to_id = to) %>%
      mutate(
        from = match(from_id, nodes$agent_id),
        to   = match(to_id,   nodes$agent_id)
      ) %>%
      select(from, to, channel, weight),
    directed = TRUE
  )

  ggraph(g, layout = "stress") +
    geom_edge_arc(
      aes(width = weight, colour = channel, alpha = weight),
      arrow     = arrow(length = unit(3, "mm"), type = "closed"),
      end_cap   = circle(6, "mm"),
      start_cap = circle(4, "mm"),
      strength  = 0.2
    ) +
    scale_edge_width(range = c(0.4, 3),  guide = "none") +
    scale_edge_alpha(range = c(0.3, 1),  guide = "none") +
    scale_edge_colour_manual(values = channel_colours, name = "Channel") +
    geom_node_point(
      aes(size = n_sent + 1, colour = colour),
      show.legend = FALSE
    ) +
    scale_colour_identity() +
    scale_size(range = c(8, 20), guide = "none") +
    geom_node_label(
      aes(label = label, colour = colour),
      repel         = TRUE,
      size          = 5,
      fontface      = "bold",
      label.padding = unit(0.3, "lines"),
      box.padding   = unit(1.5, "lines"),   # push box away from node
      point.padding = unit(1.0, "lines"),   # push away from the point itself
      force         = 5,                    # stronger repulsion
      show.legend   = FALSE
    ) +
    labs(title = plot_title, subtitle = plot_subtitle) +
    theme_graph(base_family = "sans", base_size = 14) +
    theme(
      legend.position = "bottom",
      legend.text     = element_text(size = 12),
      legend.title    = element_text(size = 13, face = "bold"),
      plot.title      = element_text(size = 15, face = "bold"),
      plot.subtitle   = element_text(size = 11, colour = "grey40")
    )
}
```


#### Network Graph to analyse crisis communication network
*Post-embargo directed network only. Shows who drove decisions
and through which channels during the crisis.*

```{r network-q1, fig.width = 9, fig.height = 7}
make_network_plot(
  period_label  = "Post-Embargo (May 23-Jun 5)",
  plot_title    = "Crisis communication network — Post-embargo period",
  plot_subtitle = paste0(
    "Directed edges = one-on-one and side-huddle messages.\n",
    "Node size = messages sent. Edge thickness = volume."
  )
)
```


::::

## 4 Evasion of embargo - step-by-step breakdown

### What this section is about

This section looks at the behaviour of the agents after the introduction of the merger embargo. We employ network, text and behavioural studies to analyse communication patterns before and after the embargo, to explore private discussions and to find discrepancies between declared and real behaviour.

### Section summary

The evidence indicates three major behavioural changes after the embargo:

* The communication went behind closed doors, with side-huddles and one-on-one conversations replacing open discussion.
* The talks went into crisis-management mode and concentrated on legal exposure, merger risks and damage control.
* There was a divergence between declared and real behaviour, with a number of agents, including Legal, engaged in activities conflicting with their stated actions.

Overall, the embargo pushed communication into less transparent routes, reducing oversight and increasing the probability of information leaking.

:::: {.panel-tabset}

### (a) Behavioural Shift — baseline vs crisis

*Pre-embargo: all communication was open comms_huddle broadcasts
(green edges — symmetric, everyone talks to everyone).
Post-embargo: bilateral side-huddle and one-on-one dominate —
Legal becomes the central hub directing others privately.*

From the network graph and the table below, we note that:

* There are significantly more conversations between the players in post-embargo. `Legal` and `Social Media` experienced the most significant jump in terms of the number of messages from 66 to 466 and 45 to 405, respectively between pre and post embargo.
* There are more players involved post embargo than one pre-embargo. `PR Intern`, `Intern`, and `Judge` are only involved post-embargo.
* Prior to onset of embargo, almost all communications are via official comms huddle. Things started to changed after embargo implementation and the players started using more private channels such as one-one-one chats and side-huddles.
* `Platform Trust` appears to be the main agent driving most of the conversations and day-to-day stewardship before the embargo implementation. Once embargo commenced, `Legal` began driving most of the conversations, as if they are in damage-control mode.

```{r network-q2, fig.width = 14, fig.height = 7}
p_pre <- make_network_plot(
  period_label  = "Pre-Embargo (May 17-22)",
  plot_title    = "Pre-Embargo: Baseline behaviour",
  plot_subtitle = "Open broadcasts only — symmetric, no hierarchy"
)

p_post <- make_network_plot(
  period_label  = "Post-Embargo (May 23-Jun 5)",
  plot_title    = "Post-Embargo: Crisis behaviour",
  plot_subtitle = "Private channels dominate — Legal as central hub"
)

(p_pre | p_post) +
  plot_annotation(
    title    = "Behavioural shift — Communication structure before vs after embargo",
    subtitle = paste0(
      "Node size = messages sent  |  ",
      "Edge thickness = volume  |  ",
      "Edge colour = channel"
    ),
    theme = theme(
      plot.title    = element_text(size = 16, face = "bold"),
      plot.subtitle = element_text(size = 12, colour = "grey40")
    )
  )
```

```{r network-q2-table}
#| echo: false

pre_deg <- network_edges %>%
  filter(period == "Pre-Embargo (May 17-22)") %>%
  group_by(from) %>%
  summarise(pre_contacts = n_distinct(to),
            pre_msgs     = sum(weight), .groups = "drop")

post_deg <- network_edges %>%
  filter(period == "Post-Embargo (May 23-Jun 5)") %>%
  group_by(from) %>%
  summarise(post_contacts = n_distinct(to),
            post_msgs     = sum(weight), .groups = "drop")

pre_deg %>%
  full_join(post_deg, by = "from") %>%
  left_join(agent_meta %>% select(agent_id, label),
            by = c("from" = "agent_id")) %>%
  mutate(
    Agent            = coalesce(label, from),
    `Pre contacts`   = coalesce(pre_contacts,  0L),
    `Post contacts`  = coalesce(post_contacts, 0L),
    `Contact change` = `Post contacts` - `Pre contacts`,
    `Pre messages`   = coalesce(pre_msgs,  0L),
    `Post messages`  = coalesce(post_msgs, 0L),
    `Msg change`     = `Post messages` - `Pre messages`
  ) %>%
  select(Agent, `Pre contacts`, `Post contacts`, `Contact change`,
         `Pre messages`, `Post messages`, `Msg change`) %>%
  arrange(desc(`Msg change`)) %>%
  knitr::kable(format = "html") %>%
  kableExtra::kable_styling(font_size = 11, full_width = FALSE) %>%
  kableExtra::row_spec(0, bold = TRUE,
                       background = "#2c3e50", color = "white")
```



### (b) Keywords before and after embargo

Here, we created two word clouds to analyse the common words before and after embargo. This can give us an idea of how the conversation (and hence behavioural changes) shifted in lieu of the embargo. 

The pre-embargo cloud shows a team doing normal business. The post-embargo cloud shows a team in legal defence mode. The word governance appears in both — but pre-embargo it meant documentation and board prep, post-embargo it meant damage control and reform commitments. Same word, completely different context — which is itself an important analytical observation about how the crisis reframed everything.

```{r wordcloud-setup}
#| include: false

pacman::p_load(tidytext, wordcloud2, htmlwidgets)

# ── custom stopwords ──────────────────────────────────────────────────────────
custom_stops <- c(
  "the", "and", "for", "this", "that", "with", "are", "have",
  "will", "from", "been", "they", "their", "what", "need", "can",
  "all", "not", "but", "its", "our", "any", "has", "was", "about",
  "more", "also", "just", "one", "out", "let", "get", "we", "it",
  "is", "in", "on", "at", "to", "of", "a", "i", "be", "do", "so",
  "if", "no", "up", "or", "an", "as", "by", "he", "she", "him",
  "her", "us", "my", "me", "re", "s", "t", "would", "should",
  "could", "make", "know", "going", "right", "team", "want", "now",
  "good", "take", "think", "time", "use", "see"
)

# ── function to build word frequency tibble ───────────────────────────────────
get_word_freq <- function(messages, min_freq = 2) {
  tibble(text = messages) %>%
    unnest_tokens(word, text) %>%
    filter(
      !word %in% custom_stops,
      !word %in% stop_words$word,
      str_length(word) > 2,
      !str_detect(word, "^\\d+$")   # remove pure numbers
    ) %>%
    count(word, sort = TRUE) %>%
    filter(n >= min_freq) %>%
    rename(freq = n)
}

# ── extract messages by period ────────────────────────────────────────────────
pre_msgs <- messages_tbl %>%
  filter(round_hour < embargo_date) %>%
  pull(content) %>%
  na.omit()

post_msgs <- messages_tbl %>%
  filter(round_hour >= embargo_date) %>%
  pull(content) %>%
  na.omit()

pre_freq  <- get_word_freq(pre_msgs)
post_freq <- get_word_freq(post_msgs)
```

```{r wordcloud-render, fig.width = 14, fig.height = 6}
#| echo: false

pacman::p_load(ggwordcloud)

# ── pre-embargo colour palette (cool blues/greens — calm) ────────────────────
pre_colours <- colorRampPalette(
  c("#1565C0", "#1D9E75", "#42A5F5", "#26C6DA", "#66BB6A")
)(nrow(pre_freq))

# ── post-embargo colour palette (warm reds/oranges — tension) ────────────────
post_colours <- colorRampPalette(
  c("#B71C1C", "#E65100", "#F57F17", "#D84315", "#C62828")
)(nrow(post_freq))

# ── pre-embargo word cloud ────────────────────────────────────────────────────
p_pre <- pre_freq %>%
  slice_max(freq, n = 80) %>%
  mutate(angle = sample(c(0, 90), n(), replace = TRUE, prob = c(0.7, 0.3))) %>%
  ggplot(aes(label = word, size = freq,
             colour = freq, angle = angle)) +
  geom_text_wordcloud(
    eccentricity = 0.65,
    seed         = 42,
    show.legend  = FALSE
  ) +
  scale_size_area(max_size = 18) +
  scale_colour_gradientn(
    colours = c("#42A5F5", "#1D9E75", "#1565C0")
  ) +
  labs(title    = "Pre-Embargo (May 17–22)",
       subtitle = "Baseline communication — planning, governance, risk") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(size = 13, face = "bold",
                                 colour = "#1565C0", hjust = 0.5),
    plot.subtitle = element_text(size = 10, colour = "grey40",
                                 hjust = 0.5),
    panel.grid    = element_blank(),
    axis.text     = element_blank()
  )

# ── post-embargo word cloud ───────────────────────────────────────────────────
p_post <- post_freq %>%
  slice_max(freq, n = 80) %>%
  mutate(angle = sample(c(0, 90), n(), replace = TRUE, prob = c(0.7, 0.3))) %>%
  ggplot(aes(label = word, size = freq,
             colour = freq, angle = angle)) +
  geom_text_wordcloud(
    eccentricity = 0.65,
    seed         = 42,
    show.legend  = FALSE
  ) +
  scale_size_area(max_size = 18) +
  scale_colour_gradientn(
    colours = c("#F57F17", "#E65100", "#B71C1C")
  ) +
  labs(title    = "Post-Embargo (May 23–Jun 5)",
       subtitle = "Crisis communication — embargo, leaks, merger, legal") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(size = 13, face = "bold",
                                 colour = "#B71C1C", hjust = 0.5),
    plot.subtitle = element_text(size = 10, colour = "grey40",
                                 hjust = 0.5),
    panel.grid    = element_blank(),
    axis.text     = element_blank()
  )

# ── side by side ─────────────────────────────────────────────────────────────
(p_pre | p_post) +
  plot_annotation(
    title    = "Word clouds — baseline vs crisis communication",
    subtitle = paste0(
      "Left: pre-embargo language (planning, governance)  |  ",
      "Right: post-embargo language (crisis, legal, merger)"
    ),
    theme = theme(
      plot.title    = element_text(size = 14, face = "bold"),
      plot.subtitle = element_text(size = 10, colour = "grey40")
    )
  )
```

### (c) Deep-dive on private chats

::: {.panel-tabset}

#### (i) data table
```{r private-convos}
#| include: false

pacman::p_load(DT)

private_tbl <- map_dfr(raw$rounds, function(r) {
  map_dfr(r$communications, function(m) {
    if (!m$channel %in% c("one_on_one_chat", "side_huddle")) return(NULL)
    tibble(
      timestamp = ymd_hms(m$timestamp),
      channel   = m$channel     %||% NA_character_,
      from      = m$agent_label %||% NA_character_,
      to        = paste(unlist(m$recipients %||% list()), collapse = ", "),
      content   = m$content     %||% NA_character_
    )
  })
}) %>%
  arrange(timestamp) %>%
  filter(
    channel == "side_huddle" |
    str_detect(
      str_to_lower(content),
      "embargo|merger|civicloom|elena|saltwind|compliance|leak|shadow|strategic development|re-identification"
    )
  ) %>%
# In private-convos chunk, change Date to sortable format:
mutate(
  Date    = format(timestamp, "%Y-%m-%d %H:%M"),   # sortable string
  Channel = if_else(channel == "one_on_one_chat", "1-on-1", "Side Huddle"),
  Period  = if_else(timestamp < embargo_date, "Pre-Embargo", "Post-Embargo"),
  To      = case_when(
    to != ""  ~ to,
    channel == "one_on_one_chat" ~ "(see context)",
    TRUE ~ "All"
  )
) %>%
select(Date, Period, Channel, From = from, To, Content = content)
# no sort_ts needed at all
```

```{=html}
<style>
  #private-conv-table tbody tr:hover {
    background-color: #eaf4fb !important;
    cursor: default;
  }
  #private-conv-table thead tr th {
    background-color: #2c3e50;
    color: white;
    padding: 8px;
    font-size: 12px;
  }
  #private-conv-table table {
    width: 100%;
    border-collapse: collapse;
    font-size: 11px;
  }
  #private-conv-table td {
    border-bottom: 1px solid #ddd;
    padding: 6px 8px;
    vertical-align: top;
  }
</style>
```

::: {#private-conv-table}
```{r private-convos-table}
#| echo: false

datatable(
  private_tbl,
  rownames = FALSE,
  filter   = "top",
  caption  = "Private conversations — one-on-one and side huddle messages",
  options  = list(
    pageLength = 5,
    lengthMenu = c(5, 10, 20, 50),
    scrollX    = TRUE,
    order      = list(list(0, "asc")),
    columnDefs = list(
      list(width = "9%",  targets = 0),
      list(width = "9%",  targets = 1),
      list(width = "9%",  targets = 2),
      list(width = "11%", targets = 3),
      list(width = "11%", targets = 4),
      list(width = "51%", targets = 5),
      list(className = "dt-center", targets = c(0, 1, 2, 3, 4))
    ),
    initComplete = JS("
      function(settings, json) {
        var api = this.api();
        [1, 2].forEach(function(colIdx) {
          var col = api.column(colIdx, {search: 'applied'});
          var select = $('<select style=\"width:100%;margin-top:4px;font-size:11px;\"><option value=\"\">All</option></select>')
            .on('change', function() {
              col.search(
                $(this).val()
                  ? '^' + $.fn.dataTable.util.escapeRegex($(this).val()) + '$'
                  : '',
                true, false
              ).draw();
            });
          col.data().unique().sort().each(function(d) {
            if (d) select.append('<option value=\"' + d + '\">' + d + '</option>');
          });
          $(col.header()).closest('thead').find('th').eq(colIdx).find('input').replaceWith(select);
        });
      }
    ")
  )
) %>%
  formatStyle(
    "Channel",
    backgroundColor = styleEqual(
      c("1-on-1",  "Side Huddle"),
      c("#EBF5FB", "#FEF9E7")
    )
  ) %>%
  formatStyle(
    "Period",
    color = styleEqual(
      c("Pre-Embargo", "Post-Embargo"),
      c("#1565C0",     "#C62828")
    ),
    fontWeight = "bold"
  ) %>%
  formatStyle(
    "Content",
    fontSize   = "11px",
    whiteSpace = "normal"
  )
```
:::

#### (ii) word clouds

##### Data Preparation
Earlier, we examined word clouds drawn from all communication channels to contrast language patterns before and after the embargo. In this section, we zoom into the private channels specifically — side huddles and one-on-one messages — to surface the behind-the-scenes conversations that were never visible to the full team.

To focus the analysis further, we introduced a set of crisis-relevant seed words (risk, embargo, leak, merger, legal, compliance, anonymous, monitor, breach, exposure) and used cosine similarity on a TF-IDF term-document matrix to identify words that were contextually associated with these seeds within the private channel corpus. Only words exceeding a similarity threshold of 0.15 were retained. This ensures the word clouds and semantic network reflect language that is analytically meaningful to the crisis — filtering out general chatter and surfacing the vocabulary agents actually used when discussing sensitive matters in private.

##### Insights from Word Clouds (based on cosine-similarity)
* The pre-embargo private channels were strikingly sparse — agents used vague, cautious language (trust, governance, window, handle) even in private, suggesting deliberate restraint in committing to specifics despite awareness that something material was coming.

* Post-embargo private language exploded in both volume and specificity. Three distinct clusters emerged: a legal/compliance cluster (governance, audit, reforms, accountability, obligation) reflecting active exposure management; a crisis actor cluster (saltwind, civicloom, oceancrunch) showing external parties had moved from background concern to front-of-mind; and a time pressure cluster (minutes, tonight, window, permanent) signalling real-time crisis execution rather than planned communication.

* The contrast is analytically significant — the sparseness of pre-embargo private language suggests the team knew more than they were saying even in private, while the post-embargo density reveals the crisis forced specificity that careful embargo management had been suppressing.

```{r cosine-setup}
pacman::p_load(text2vec, tidytext, ggraph, tidygraph, ggwordcloud, DT)

# ── seed words ────────────────────────────────────────────────────────────────
seed_words <- c(
  "risk", "embargo", "leak", "merger", "legal",
  "compliance", "anonymous", "monitor", "breach", "exposure"
)

# ── stopwords ─────────────────────────────────────────────────────────────────
extra_stops <- c(
  stop_words$word,
  "the","and","for","this","that","with","are","have","will",
  "from","been","they","their","what","need","can","all","not",
  "but","its","our","any","has","was","re","s","t","it","is",
  "in","on","at","to","of","a","i","be","do","so","if","no",
  "up","or","an","as","by","we","he","she","him","her","us",
  "my","me","would","should","could","make","know","going",
  "right","team","want","now","good","take","think","time",
  "use","see","just","also","one","get","let","out","more",
  "about","very","here","there","when","then","them","these",
  "those","into","even","both","back","well","still",
  "platform_trust","social_manager","legal_agent",
  "pr_agent","pr_intern","judge_agent","intern_agent"
)

# ── prepare corpus from private_tbl ──────────────────────────────────────────
corpus <- private_tbl %>%
  mutate(
    message_id = as.character(row_number()),
    content    = str_to_lower(Content) %>%
                 str_replace_all("[^a-z\\s]", " ") %>%
                 str_squish()
  ) %>%
  select(message_id, content) %>%
  deframe()

# ── build vocabulary ──────────────────────────────────────────────────────────
it <- itoken(corpus, tokenizer = word_tokenizer, progressbar = FALSE)

vocab <- create_vocabulary(it) %>%
  prune_vocabulary(term_count_min = 2, doc_count_min = 1) %>%
  filter(!term %in% extra_stops, nchar(term) > 2)

vectoriser  <- vocab_vectorizer(vocab)
dtm         <- create_dtm(it, vectoriser)
tfidf       <- TfIdf$new()
dtm_tfidf   <- fit_transform(dtm, tfidf)
tdm         <- Matrix::t(dtm_tfidf)

# ── cosine similarity ─────────────────────────────────────────────────────────
similarity_threshold <- 0.15

sim_results <- map_dfr(seed_words, function(seed) {
  if (!seed %in% rownames(tdm)) return(NULL)
  seed_vec <- tdm[seed, , drop = FALSE]
  all_sims <- sim2(tdm, seed_vec, method = "cosine", norm = "l2")
  tibble(
    seed       = seed,
    word       = rownames(all_sims),
    similarity = as.numeric(all_sims)
  ) %>%
    filter(
      word != seed,
      !word %in% seed_words,
      !word %in% extra_stops,
      similarity > similarity_threshold
    ) %>%
    slice_max(similarity, n = 12)
})

# ── seed-associated vocabulary for word clouds ────────────────────────────────
seed_vocab <- sim_results %>% pull(word) %>% unique()

# ── word frequency filtered to seed vocab ────────────────────────────────────
get_seed_freq <- function(period_filter) {
  private_tbl %>%
    filter(Period == period_filter) %>%
    mutate(message_id = row_number()) %>%
    unnest_tokens(word, Content) %>%
    filter(
      word %in% seed_vocab,
      !word %in% extra_stops,
      str_length(word) > 2
    ) %>%
    count(word, sort = TRUE) %>%
    rename(freq = n)
}

pre_freq  <- get_seed_freq("Pre-Embargo")
post_freq <- get_seed_freq("Post-Embargo")

# ── network graph ─────────────────────────────────────────────────────────────
cosine_edges <- sim_results %>%
  rename(from = seed, to = word, weight = similarity)

network_nodes <- tibble(
  name    = unique(c(cosine_edges$from, cosine_edges$to)),
  is_seed = name %in% seed_words
)

cosine_graph <- cosine_edges %>%
  as_tbl_graph() %>%
  activate(nodes) %>%
  left_join(network_nodes, by = "name") %>%
  mutate(is_seed = coalesce(is_seed, FALSE))
```

```{r cosine-wordclouds, fig.width = 14, fig.height = 6}
#| echo: false

p_pre <- pre_freq %>%
  slice_max(freq, n = 50) %>%
  mutate(angle = sample(c(0, 90), n(), replace = TRUE,
                        prob = c(0.7, 0.3))) %>%
  ggplot(aes(label = word, size = freq,
             colour = freq, angle = angle)) +
  geom_text_wordcloud(eccentricity = 0.65, seed = 42) +
  scale_size_area(max_size = 16) +
  scale_colour_gradientn(
    colours = c("#42A5F5", "#1D9E75", "#1565C0")
  ) +
  labs(
    title    = "Pre-Embargo (May 17–22)",
    subtitle = "Seed-associated words only"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 13, face = "bold",
                                 colour = "#1565C0", hjust = 0.5),
    plot.subtitle = element_text(size = 9, colour = "grey40",
                                 hjust = 0.5),
    panel.grid    = element_blank(),
    axis.text     = element_blank()
  )

p_post <- post_freq %>%
  slice_max(freq, n = 50) %>%
  mutate(angle = sample(c(0, 90), n(), replace = TRUE,
                        prob = c(0.7, 0.3))) %>%
  ggplot(aes(label = word, size = freq,
             colour = freq, angle = angle)) +
  geom_text_wordcloud(eccentricity = 0.65, seed = 42) +
  scale_size_area(max_size = 16) +
  scale_colour_gradientn(
    colours = c("#F57F17", "#E65100", "#B71C1C")
  ) +
  labs(
    title    = "Post-Embargo (May 23–Jun 5)",
    subtitle = "Seed-associated words only"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 13, face = "bold",
                                 colour = "#B71C1C", hjust = 0.5),
    plot.subtitle = element_text(size = 9, colour = "grey40",
                                 hjust = 0.5),
    panel.grid    = element_blank(),
    axis.text     = element_blank()
  )

(p_pre | p_post) +
  plot_annotation(
    title    = "Word clouds — seed-associated terms (private channels)",
    subtitle = paste0(
      "Only words with cosine similarity > ", similarity_threshold,
      " to at least one seed word are shown.\n",
      "Word size = frequency in private messages."
    ),
    theme = theme(
      plot.title    = element_text(size = 14, face = "bold"),
      plot.subtitle = element_text(size = 10, colour = "grey40")
    )
  )
```


#### (iii) semantic network

The **semantic network** visualises cosine similarity scores between seed words and their associated terms. Nodes connected by thicker edges share stronger contextual similarity in the private channel corpus. Only words exceeding the 0.15 threshold are shown, keeping the network focused on analytically relevant associations.

The top three most interesting clusters and insights gathered are:

* **breach cluster** — the words `elenamarquez`, `accidentally`, `justification`, `mac`, and `survival` reconstruct the @Elena incident in a single cluster. `accidentally` and `justification` suggest agents were privately framing the breach as unintentional, while `mac` (which stands for Material Adverse Change) and `survival` reveal they understood the real stakes — the breach had potentially handed CivicLoom a Material Adverse Change clause exit from the merger without penalty.

* **legal cluster** — `civicloom`, `saltwind`, `oceancrunch`, `minutes`, and `governance` all orbit `legal`, confirming Legal was simultaneously managing all external parties — press, merger counterparty, and enterprise clients — from a single coordination point.

* **risk cluster** — `incident`, `inquiries`, `calculus`, `shrinking`, `investment`, `revenue`, and `contained` surround `risk`. The words `shrinking` and `revenue` confirm agents were privately connecting the crisis to financial fundamentals, not just reputational damage.  The word `calculus` is particularly telling — agents were not reacting emotionally to risk but performing deliberate cost-benefit reasoning. `inquiries` links back to the AG investigations. The word calculus is the most telling – it comes from CEO Ajay’s private language: “the risk calculus changes significantly”. The agents were not utilising general risk terminology but instead were echoing the CEO’s own framing, implying that the private risk reasoning of the entire team was being shaped top-down by signals from Ajay.

```{r cosine-network-plot, fig.width = 14, fig.height = 9}
#| echo: false

ggraph(cosine_graph, layout = "fr") +
  geom_edge_link(
    aes(width = weight, alpha = weight),
    colour      = "grey55",
    show.legend = FALSE
  ) +
  scale_edge_width(range  = c(0.2, 3),    guide = "none") +
  scale_edge_alpha(range  = c(0.15, 0.9), guide = "none") +
  geom_node_point(
    data   = . %>% filter(!is_seed),
    colour = "#3266ad", size = 4, alpha = 0.75
  ) +
  geom_node_point(
    data   = . %>% filter(is_seed),
    colour = "#B71C1C", size = 10, alpha = 0.9
  ) +
  geom_node_label(
    aes(
      label    = name,
      colour   = if_else(is_seed, "#B71C1C", "#1565C0"),
      fontface = if_else(is_seed, "bold", "plain")
    ),
    size          = 3.5,
    repel         = TRUE,
    label.padding = unit(0.25, "lines"),
    show.legend   = FALSE
  ) +
  scale_colour_identity() +
  labs(
    title    = "Semantic network — cosine similarity (private channels)",
    subtitle = paste0(
      "Red = seed words  |  Blue = associated terms\n",
      "Edge thickness = cosine similarity  |  ",
      "Threshold = ", similarity_threshold
    )
  ) +
  theme_graph(base_family = "sans", base_size = 12) +
  theme(
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10, colour = "grey40")
  )
```

#### (iv) TF-IDF

While cosine similarity identifies words contextually associated with crisis seed terms, **TF-IDF ((Term Frequency-Inverse Document Frequency)** complements this by revealing what is distinctive about each agent's private vocabulary relative to others. 

By clicking on the 'Cosine similarity' column to order it in descending order. The four strongest associations are: `thinner` (0.882), `logged` (0.881), `noted` (0.879), creates (0.831). 

It is worth noting that `monitor` carries several distinct meanings in this corpus, which the cosine similarity analysis captures in aggregate:

* **Compliance monitor** — the Judge agent assigned post-breach on May 30 (For information, 'compliance monitor' actually refers to the `judge_agent` as evident in the 1-on-1 content on May 29 in the data table below.)
* **Risk monitoring** — Platform Trust's internal risk score tracking
* **Real-time social surveillance** — Crisis Day monitoring of ResidentIQ IR, @PropTechWatcher, and FTC advocacy groups

Taken together, it means that these words appeared in the most similar message contexts as the word `monitor` across the private channel corpus. In plain terms — whenever agents privately discussed the monitor, they almost always did so in the context of documentation, accountability, and narrowing margins. 


```{r cosine-top-words}
#| echo: false

sim_results %>%
  group_by(seed) %>%
  slice_max(similarity, n = 5) %>%
  ungroup() %>%
  mutate(similarity = round(similarity, 3)) %>%
  rename(
    `Seed word`         = seed,
    `Associated word`   = word,
    `Cosine similarity` = similarity
  ) %>%
  datatable(
    rownames = FALSE,
    filter   = "top",
    options  = list(
      pageLength = 10,
      scrollX    = TRUE,
      autoWidth  = FALSE,
      columnDefs = list(
        list(width = "150px", targets = 0),
        list(width = "150px", targets = 1),
        list(width = "150px", targets = 2)
      )
    ),
    caption = "Top 5 words associated with each seed — cosine similarity"
  ) %>%
  formatStyle(
    "Cosine similarity",
    background         = styleColorBar(c(0, 1), "#BBDEFB"),
    backgroundSize     = "100% 90%",
    backgroundRepeat   = "no-repeat",
    backgroundPosition = "center"
  ) %>%
  formatStyle(
    "Seed word",
    fontWeight = "bold",
    color      = "#B71C1C"
  )
```

```{r monitor-evidence}
#| echo: false

private_tbl %>%
  filter(
    str_detect(
      str_to_lower(Content),
      "monitor|logged|noted|thinner|creates|minimum"
    )
  ) %>%
  mutate(
    matched = str_extract_all(
      str_to_lower(Content),
      "monitor|logged|noted|thinner|creates|minimum"
    ) %>% map_chr(~paste(unique(.x), collapse = ", "))
  ) %>%
  select(Date, Period, Channel, From, Content, matched) %>%
  datatable(
    rownames = FALSE,
    filter   = "top",
    options  = list(
      pageLength = 5,
      scrollX    = TRUE,
      autoWidth  = FALSE,
      initComplete = JS("
        function(settings, json) {
          var api = this.api();
          $(api.table().node()).css('font-size', '11px');
          api.columns(0).nodes().to$().css('min-width', '80px').css('max-width', '80px');
          api.columns(1).nodes().to$().css('min-width', '80px').css('max-width', '80px');
          api.columns(2).nodes().to$().css('min-width', '80px').css('max-width', '80px');
          api.columns(3).nodes().to$().css('min-width', '80px').css('max-width', '80px');
          api.columns(4).nodes().to$().css('min-width', '350px');
          api.columns(5).nodes().to$().css('min-width', '120px').css('max-width', '120px');
          api.columns.adjust().draw();
        }
      ")
    ),
    caption = "Private messages containing monitor-related terms"
  ) %>%
  formatStyle(
    "matched",
    fontWeight = "bold",
    color      = "#B71C1C"
  )
```

:::

### (d) Declared vs actual behaviour

Below is a swimlane plot created by joining `messages_tbl` (actual channel usage) with the `participants` field (declared actions). The red border means that the agent declared `monitoring` officially, but their actual channel usage shows that they were actually posting on other channels (`anonymous_post`, `personal_post`, or `side_huddle`.)

From the swimlane plot, we note the following:

* **Pre-embargo** — all agents show clean monitoring (light blue) with only official posts (dark blue). Legal's yellow tile on May 22 — side huddle only — is the first signal of off-record activity, one day before the embargo.
* **Intern-Agent anomaly**  — personal posting (orange) continuously from May 24 through June 1 despite being explicitly excluded from the merger briefing. Posting without awareness of the risk — an uninformed actor operating in an informed crisis.
* **Social Media and Legal — key divergences on Crisis Day** — both show red borders (declared MONITORING, actual personal/anonymous posting). Legal's dark red tiles are the most analytically significant — the agent responsible for compliance was itself behaving inconsistently with its declared actions.
* **Judge-Agent** — one grey tile at Jun 5 15:00, otherwise absent. A single compliance warning across the entire observation period confirms the monitor was observational, not preventive.
* **Platform-Trust-Agent** — clean throughout until Crisis Day afternoon when red borders appear, suggesting even the most disciplined agent shifted under crisis pressure.

```{r declared-actual-setup}
#| include: false

# ── Use actual channel behaviour as proxy throughout all 23 rounds ────────────
action_tbl <- messages_tbl %>%
  group_by(round_hour, agent_id, agent_label) %>%
  summarise(
    channels   = list(unique(channel)),
    n_messages = n(),
    .groups    = "drop"
  ) %>%
  mutate(
    effective_action = case_when(
      map_lgl(channels, ~any(.x == "anonymous_post")) ~ "Anonymous post",
      map_lgl(channels, ~any(.x == "personal_post"))  ~ "Personal post",
      map_lgl(channels, ~any(.x == "official_post"))  ~ "Official post",
      map_lgl(channels, ~any(.x == "side_huddle"))    ~ "Side huddle only",
      TRUE                                             ~ "Monitoring"
    ),
    period = case_when(
      round_hour < embargo_date                          ~ "Pre-Embargo",
      round_hour >= embargo_date &
        as.Date(round_hour) < as.Date("2046-06-05")     ~ "Post-Embargo",
      TRUE                                               ~ "Crisis Day"
    ),
    agent_label = coalesce(agent_label, agent_id)
  )

# ── Extract Crisis Day declared actions from participants ─────────────────────
declared_crisis <- map_dfr(raw$rounds, function(r) {
  map_dfr(r$participants %||% list(), function(p) {
    da <- p$declared_action %||% NA_character_
    if (is.na(da)) return(NULL)
    tibble(
      round_hour    = ymd_hms(r$hour),
      agent_id      = p$agent_id %||% NA_character_,
      declared      = da,
      declared_type = case_when(
        str_detect(da, "POSTED_ANONYMOUS")   ~ "Anonymous post",
        str_detect(da, "POSTED_PERSONAL")    ~ "Personal post",
        str_detect(da, "POSTED_ON_FLEX")     ~ "Official post",
        str_detect(da, "COMPLIANCE_WARNING") ~ "Compliance warning",
        str_detect(da, "MONITORING")         ~ "Monitoring",
        TRUE                                 ~ "Other"
      )
    )
  })
})

# ── Merge and build final display table ──────────────────────────────────────
action_final_tbl <- action_tbl %>%
  left_join(
    declared_crisis %>% select(round_hour, agent_id, declared_type),
    by = c("round_hour", "agent_id")
  ) %>%
  mutate(
    display_action = coalesce(declared_type, effective_action),
    display_action = factor(display_action,
      levels = c("Monitoring", "Side huddle only", "Official post",
                 "Personal post", "Anonymous post",
                 "Compliance warning", "Other")
    ),
    divergence = case_when(
      !is.na(declared_type) &
        declared_type == "Monitoring" &
        effective_action %in% c("Anonymous post", "Personal post",
                                "Side huddle only") ~ "DIVERGED",
      TRUE ~ "Consistent"
    ),
    round_label = format(round_hour, "%b %d %H:%M"),
    round_label = factor(round_label,
                         levels = unique(round_label[order(round_hour)]))
  )

# ── Colour palette ────────────────────────────────────────────────────────────
action_colours <- c(
  "Monitoring"         = "#BDE0F5",
  "Side huddle only"   = "#FFF3CD",
  "Official post"      = "#1565C0",
  "Personal post"      = "#E65100",
  "Anonymous post"     = "#C62828",
  "Compliance warning" = "#5F5E5A",
  "Other"              = "#BBBBBB"
)

# ── Period boundary labels for annotation ────────────────────────────────────
period_labels <- action_final_tbl %>%
  group_by(period) %>%
  slice_min(round_hour, n = 1) %>%
  ungroup() %>%
  distinct(period, round_label) %>%
  mutate(
    colour = case_when(
      period == "Pre-Embargo"  ~ "#1565C0",
      period == "Post-Embargo" ~ "#E65100",
      period == "Crisis Day"   ~ "#B71C1C"
    ),
    y_pos = length(unique(action_final_tbl$agent_label)) + 0.7
  )
```

```{r declared-actual-heatmap, fig.width = 14, fig.height = 6}

n_agents <- length(unique(action_final_tbl$agent_label))

ggplot(action_final_tbl,
       aes(x = round_label, y = agent_label, fill = display_action)) +

  # main tiles
  geom_tile(colour = "white", linewidth = 0.5) +

  # red border on diverged rounds
  geom_tile(
    data      = action_final_tbl %>% filter(divergence == "DIVERGED"),
    aes(x     = round_label, y = agent_label),
    fill      = NA,
    colour    = "#B71C1C",
    linewidth = 1.8
  ) +

  ## ----------add red line to mark embargo date -----------
geom_vline(
  xintercept = which(levels(action_final_tbl$round_label) == 
                       format(embargo_date, "%b %d %H:%M")),
  colour    = "#B71C1C",
  linewidth = 1.2,
  linetype  = "solid"
) +
annotate(
  "text",
  x      = which(levels(action_final_tbl$round_label) == 
                   format(embargo_date, "%b %d %H:%M")) + 0.15,
  y      = 2.0,
  label  = "Embargo declared",
  colour = "#B71C1C",
  size   = 3,
  hjust  = 0,
  fontface = "italic"
) +

  ## ----------end of embargo date chunk -------------------
  
  
  # period annotation labels
  geom_text(
    data     = period_labels,
    aes(x    = round_label, y = y_pos,
        label = period, colour = colour),
    hjust    = 0,
    size     = 3.5,
    fontface = "bold",
    inherit.aes = FALSE
  ) +

  scale_fill_manual(values = action_colours, name = "Action") +
  scale_colour_identity() +
  scale_y_discrete(expand = expansion(add = c(0.5, 1.2))) +

  labs(
    title    = "Agent behaviour across all 23 rounds — Pre-Embargo to Crisis Day",
    subtitle = paste0(
  "Colour = effective action (actual channels used). White = agent not active that round.\n",
  "On Crisis Day, declared action shown where available.\n",
  "Red border = declared MONITORING but used anonymous/personal/side-huddle channels.\n",
  "Red vertical line = embargo declared (May 23)."
    ),
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y     = element_text(size = 10),
    legend.position = "bottom",
    plot.title      = element_text(size = 13, face = "bold"),
    plot.subtitle   = element_text(size = 9,  colour = "grey40"),
    panel.grid      = element_blank()
  )
```

```{r declared-actual-diverge-table}
#| echo: false

diverged_tbl <- action_final_tbl %>%
  filter(divergence == "DIVERGED") %>%
  mutate(
    Date             = format(round_hour, "%d %b %H:%M"),
    Agent            = agent_label,
    `Declared`       = coalesce(declared_type, "—"),
    `Actual channels`= map_chr(channels, ~paste(sort(.x), collapse = ", ")),
    `Messages sent`  = n_messages
  ) %>%
  select(Date, Agent, Declared, `Actual channels`, `Messages sent`)
```

```{=html}
<style>
  #diverge-table tbody tr:hover {
    background-color: #fdecea !important;
    cursor: default;
  }
  #diverge-table thead tr th {
    background-color: #B71C1C;
    color: white;
    padding: 8px;
    font-size: 12px;
  }
  #diverge-table table {
    width: 100%;
    border-collapse: collapse;
    font-size: 11px;
  }
  #diverge-table td {
    border-bottom: 1px solid #ddd;
    padding: 6px 8px;
    vertical-align: top;
  }
  #diverge-table table td:nth-child(1) { width: 10%; white-space: nowrap; }
  #diverge-table table td:nth-child(2) { width: 12%; white-space: nowrap; }
  #diverge-table table td:nth-child(3) { width: 15%; white-space: nowrap; }
  #diverge-table table td:nth-child(4) { width: 48%; white-space: normal; }
  #diverge-table table td:nth-child(5) { width: 10%; text-align: center;  }
</style>
```

*Rounds where agent declared **MONITORING** but actually used anonymous, personal, or side-huddle channels:*

::: {#diverge-table}
```{r diverge-table-render}
#| echo: false
diverged_tbl %>%
  knitr::kable(format = "html") %>%
  kableExtra::kable_styling(font_size = 11, full_width = TRUE) %>%
  kableExtra::row_spec(0, bold = TRUE,
                       background = "#B71C1C", color = "white")
```
:::


::::



## 5 Leading indicators of inappropriate release

### What this section is about

This section details the leading indicators that preceded the improper disclosure of information about the mergers. We look at internal risk signals, warning messages and posting behaviour to identify agents who were causing risk, and those who were actively flagging it.

### Section summary

There were warning indicators present before the breach occurred:

* **The risk score of Platform Trust rose gradually** and reached a high level one day before the breach.
* **Legal, Platform Trust, Social Media and PR raised concerns privately** regularly but these warnings never made it up the chain successfully.
* **After the embargo, personal posting conduct was established and normalised** among several agents, without assistance.
* **Risk makers and risk flaggers communicated via different channels of communication**, creating a governance gap where warning signals were not impacting conduct.

There were plenty of warning signs overall that could have been detected prior to the breach, but the organisation lacked an efficient way to tie risk identification to corrective action.

```{r leading-indicators-setup}
#| include: false

# ── 1. Personal posts BEFORE breach (habit formation) ────────────────────────
pre_personal_posts <- messages_tbl %>%
  filter(
    channel %in% c("personal_post", "anonymous_post"),
    round_hour < ymd_hms("2046-05-29T09:00:00")
  ) %>%
  mutate(
    signal_type = "Personal/anonymous post",
    label       = paste0(agent_label, "\n", str_trunc(content, 60))
  )

# ── 2. Risk scores from deliberating BEFORE breach ───────────────────────────
pre_risk_scores <- map_dfr(raw$rounds, function(r) {
  if (ymd_hms(r$hour) >= ymd_hms("2046-05-29T09:00:00")) return(NULL)
  map_dfr(r$communications, function(m) {
    ist   <- m$internal_state %||% list()
    delib <- ist$deliberating %||% ""
    scores <- str_extract_all(delib, "\\d+\\.?\\d*(?=\\s*/\\s*10)")[[1]]
    if (length(scores) == 0) return(NULL)
    tibble(
      timestamp  = ymd_hms(m$timestamp),
      agent      = m$agent_label %||% NA_character_,
      risk_score = as.numeric(scores[1]),
      context    = delib #str_trunc(delib, 120)
    )
  })
}) %>%
  filter(!is.na(risk_score), risk_score <= 10)

# ── 3. Warning signals from deliberating BEFORE breach ───────────────────────
pre_warnings <- map_dfr(raw$rounds, function(r) {
  if (ymd_hms(r$hour) >= ymd_hms("2046-05-29T09:00:00")) return(NULL)
  map_dfr(r$communications, function(m) {
    ist   <- m$internal_state %||% list()
    delib <- (ist$deliberating %||% "")
    if (!str_detect(str_to_lower(delib),
                    "concern|worry|risk|warning|one bad|narrative|hard to manage|compounding")) {
      return(NULL)
    }
    tibble(
      timestamp   = ymd_hms(m$timestamp),
      agent       = m$agent_label %||% NA_character_,
      signal_type = "Warning in deliberating",
      excerpt     = delib #str_trunc(delib, 100)
    )
  })
})

# ── breach event marker and elena post time ───────────────────────────────────
breach_time     <- ymd_hms("2046-05-29T09:11:00")
elena_post_time <- ymd_hms("2046-05-29T09:11:00")

# ── judge first appearance ────────────────────────────────────────────────────
judge_start <- messages_tbl %>%
  filter(agent_id == "judge_agent") %>%
  summarise(first = min(round_hour)) %>%
  pull(first)

# ── for q3b-informal-posts chunk ─────────────────────────────────────────────
informal_posts_tbl <- messages_tbl %>%
  filter(channel %in% c("personal_post", "anonymous_post")) %>%
  mutate(
    period = case_when(
      round_hour <  embargo_date                     ~ "Pre-Embargo",
      round_hour >= embargo_date &
        as.Date(round_hour) < as.Date(crisis_date)  ~ "Post-Embargo",
      TRUE                                           ~ "Crisis Day"
    ),
    is_risky = str_detect(
      str_to_lower(coalesce(content, "")),
      "big things|strategic|civicloom|merger|harborcrest|embargo|elena|exciting|interesting week"
    ),
    risk_label = if_else(is_risky, "Risky content", "Benign content"),
    Date       = format(round_hour, "%d %b %H:%M")
  ) %>%
  arrange(round_hour)
```

:::: {.panel-tabset}
### (a) Leading indicators plot

The Platform Trust agent tracks an internal risk score within its private deliberating field, expressed as a value out of 10. This score is extracted using a regex pattern that identifies values in the form x/10. As the only agent that consistently recorded a numerical self-assessment, the score reflects its own perception of the team's accumulating exposure rather than an externally defined measure. 

Here, we visualise these scores in a scatter plot over the two-week period to examine risk trends and identify potential leading signals preceding key events. The following insights can be gleaned from the plot:

* **The only substantive pre-breach risk signal** - is the data pertainint to Platform Trust's risk score made on May 28, one day before the breach. Its appearance just after the embargo suggests that formal risk tracking began too late.
* **Warning signs (diamonds)** – emerge from May 17 across Legal, Platform Trust, Social Media and PR, rising in frequency post embargo. Several agents were raising concerns privately at the same time but no alert ever appeared in a formal way or made its way to the agents whose behaviour needed to alter.
* **Personal posts (triangles)** - Only appear after May 23, 3 agents develop a posting habit in the 6-day period between embargo and violation. No interference, no correction, no penalties. 
* **The essential gap** — alert signals pile up at one y-axis level, personal messages at another. Different agents, no connection. The visual gap between the two rows is the governance gap – the agents signalling risk and the agents generating it never met.


#### scatter plot on leading indicator risks

```{r leading-indicators-plot, fig.width = 12, fig.height = 7}
#| echo: false

# ── colour mapping ────────────────────────────────────────────────────────────
agent_cols <- c(
  "Legal-Agent"          = "#3266ad",
  "Platform-Trust-Agent" = "#1D9E75",
  "Social-Manager-Agent" = "#D85A30",
  "PR-Agent"             = "#B07EB8",
  "PR-Intern-Agent"      = "#BA7517",
  "Intern-Agent"         = "#888780"
)

ggplot() +

  # ── background shading ────────────────────────────────────────────────────
  annotate("rect",
           xmin  = as.numeric(ymd_hms("2046-05-17T00:00:00")),
           xmax  = as.numeric(embargo_date),
           ymin  = -Inf, ymax = Inf,
           fill  = "#E3F2FD", alpha = 0.3) +
  annotate("rect",
           xmin  = as.numeric(embargo_date),
           xmax  = as.numeric(breach_time),
           ymin  = -Inf, ymax = Inf,
           fill  = "#FFF3E0", alpha = 0.3) +

  # ── risk score line ───────────────────────────────────────────────────────
  geom_line(
    data    = pre_risk_scores,
    aes(x   = timestamp, y = risk_score * 0.8),
    colour  = "#1D9E75", linewidth = 1.2, linetype = "dashed"
  ) +
  geom_point(
    data    = pre_risk_scores,
    aes(x   = timestamp, y = risk_score * 0.8),
    colour  = "#1D9E75", size = 3
  ) +
  geom_label(
    data    = pre_risk_scores,
    aes(x   = timestamp, y = risk_score * 0.8 + 0.3,
        label = paste0(agent, "\nRisk: ", risk_score, "/10")),
    size    = 2.8, colour = "#1D9E75",
    label.padding = unit(0.2, "lines"), fill = "white"
  ) +

  # ── warning signals ───────────────────────────────────────────────────────
  geom_point(
    data    = pre_warnings,
    aes(x   = timestamp, y = 2.5, colour = agent),
    shape   = 18, size = 5, alpha = 0.8
  ) +
ggrepel::geom_text_repel(
  data    = pre_warnings,
  aes(x   = timestamp, y = 2.5, label = agent, colour = agent),
  size         = 2.8,
  max.overlaps = 20,
  box.padding  = 0.4,
  show.legend  = FALSE
) +
  # ── personal posts ────────────────────────────────────────────────────────
  geom_point(
    data    = pre_personal_posts,
    aes(x   = round_hour, y = 1.0, colour = agent_label),
    shape   = 17, size = 4, alpha = 0.8
  ) +
ggrepel::geom_text_repel(
  data    = pre_personal_posts %>% distinct(round_hour, agent_label),
  aes(x   = round_hour, y = 1.0, label = agent_label, colour = agent_label),
  size         = 2.8,
  max.overlaps = 20,
  box.padding  = 0.4,
  show.legend  = FALSE
) +

  # ── embargo line ──────────────────────────────────────────────────────────
  geom_vline(xintercept = as.numeric(embargo_date),
             colour = "#E65100", linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = embargo_date, y = 6.5,
           label = "Embargo\n(May 23)", hjust = -0.1,
           size = 3, colour = "#E65100", fontface = "italic") +

  # ── breach line ───────────────────────────────────────────────────────────
  geom_vline(xintercept = as.numeric(breach_time),
             colour = "#B71C1C", linewidth = 1.2) +
  annotate("text", x = breach_time, y = 6.5,
           label = "@Elena post\n(May 29)", hjust = -0.1,
           size = 3, colour = "#B71C1C", fontface = "bold") +

  # ── row labels ────────────────────────────────────────────────────────────
 # annotate("text", x = ymd_hms("2046-05-17T12:00:00"), y = 5.2,
 #         label = "Risk score (/10)", hjust = 0,
 #           size = 3, colour = "#1D9E75", fontface = "bold") +
  # annotate("text", x = ymd_hms("2046-05-17T12:00:00"), y = 2.7,
  #          label = "Warning signals\n(deliberating)", hjust = 0,
  #          size = 3, colour = "#555555", fontface = "bold") +
  # annotate("text", x = ymd_hms("2046-05-17T12:00:00"), y = 1.2,
  #          label = "Personal posts\n(habit forming)", hjust = 0,
  #          size = 3, colour = "#555555", fontface = "bold") +

  scale_colour_manual(values = agent_cols, name = "Agent") +
  scale_x_datetime(date_labels = "%b %d", date_breaks = "3 days") +
  scale_y_continuous(
    name   = "Risk score (/10)",
    limits = c(0, 7)
  ) +

  labs(
    title    = "Leading indicators — pre-breach signals (May 17–28)",
    subtitle = paste0(
      "Three converging signals before the @Elena post on May 29:\n",
      "Rising internal risk score (green dashed) | ",
      "Warning signals in private deliberating (diamonds) | ",
      "Personal posting habit forming (triangles)"
    ),
    x = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1),
    legend.position  = "bottom",
    plot.title       = element_text(size = 13, face = "bold"),
    plot.subtitle    = element_text(size = 9, colour = "grey40"),
    panel.grid.minor = element_blank()
  )
```

#### Leading indicators table 

```{r leading-indicators-table}
#| echo: false

bind_rows(
  pre_personal_posts %>%
    transmute(
      timestamp,
      agent    = agent_label,
      signal   = "Personal/anonymous post",
      evidence = content #str_trunc(content, 200)
    ),
  pre_warnings %>%
    transmute(
      timestamp,
      agent    = agent,
      signal   = "Warning in deliberating",
      evidence = excerpt
    ),
  pre_risk_scores %>%
    transmute(
      timestamp,
      agent    = agent,
      signal   = paste0("Risk score: ", risk_score, "/10"),
      evidence = context
    )
) %>%
  arrange(timestamp) %>%
  mutate(Date = format(timestamp, "%d %b %H:%M")) %>%
  select(Date, Agent = agent, Signal = signal, Evidence = evidence) %>%
  datatable(
    rownames = FALSE,
    filter   = "top",
    options  = list(
      pageLength = 5,
      scrollX    = TRUE,
       scrollY    = "400px",  
      autoWidth  = FALSE,
      initComplete = JS("
        function(settings, json) {
          var api = this.api();
          $(api.table().node()).css('font-size', '11px');
          api.columns(0).nodes().to$().css('min-width', '90px').css('max-width', '90px');
          api.columns(1).nodes().to$().css('min-width', '110px').css('max-width', '110px');
          api.columns(2).nodes().to$().css('min-width', '150px').css('max-width', '150px');
          api.columns(3).nodes().to$().css('min-width', '600px');
          api.columns.adjust().draw();
        }
      ")
    ),
    caption = "Pre-breach leading indicator signals — May 17 to May 28"
  ) %>%
  formatStyle(
    "Signal",
    color = styleEqual(
      c("Personal/anonymous post", "Warning in deliberating"),
      c("#E65100", "#3266ad")
    ),
    fontWeight = "bold"
  )
```

### (b) Risk creators vs Risk Flaggers

**Panel 1 identifies the risk creators**: 

* Social Media agent posted personally on May 23, the same day the embargo was informally declared, establishing the first instance of the behaviour. 
* PR Intern and Intern followed from May 24 onwards, each posting company-positive content on personal accounts. 
* By May 28 — one day before the breach — three agents had normalised personal posting with no consequences, no correction, and no awareness that an embargo was in force.

**Panel 2 identifies the risk flaggers**:

* Legal and Platform Trust were privately raising concerns from May 17 onwards, with warnings appearing in their deliberating field across every pre-breach round. 
* Platform Trust's risk score climbed from 4.2/10 on May 17 to 5.8/10 on May 28 — already in the elevated zone before the breach occurred. 
* Social Media and PR also appear as flaggers, indicating even they privately sensed the risk environment was tightening.

```{r risk-split-plot, fig.width = 13, fig.height = 9}
#| echo: false

pacman::p_load(ggrepel, patchwork)

# ── Panel 1: Risk CREATORS — agents posting personally pre-breach ─────────────
p_creators <- ggplot() +

  annotate("rect",
           xmin = as.numeric(ymd_hms("2046-05-17T00:00:00")),
           xmax = as.numeric(embargo_date),
           ymin = -Inf, ymax = Inf,
           fill = "#E3F2FD", alpha = 0.4) +
  annotate("rect",
           xmin = as.numeric(embargo_date),
           xmax = as.numeric(breach_time),
           ymin = -Inf, ymax = Inf,
           fill = "#FFF3E0", alpha = 0.4) +

  # personal post points
  geom_point(
    data  = pre_personal_posts,
    aes(x = round_hour, y = agent_label, colour = agent_label),
    shape = 17, size = 5, alpha = 0.85
  ) +

  # count label per agent per round
  geom_label(
    data  = pre_personal_posts %>%
              count(round_hour, agent_label),
    aes(x = round_hour, y = agent_label,
        label = paste0("n=", n)),
    size  = 3.5, nudge_y = 0.35,
    label.padding = unit(0.15, "lines"),
    show.legend = FALSE
  ) +

  # embargo and breach lines
  geom_vline(xintercept = as.numeric(embargo_date),
             colour = "#E65100", linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = embargo_date,
           y = length(unique(pre_personal_posts$agent_label)) + 0.5,
           label = "Embargo (May 23)",
           hjust = -0.1, size = 3,
           colour = "#E65100", fontface = "italic") +

  geom_vline(xintercept = as.numeric(breach_time),
             colour = "#B71C1C", linewidth = 1.2) +
  annotate("text", x = breach_time,
           y = length(unique(pre_personal_posts$agent_label)) + 0.5,
           label = "@Elena (May 29)",
           hjust = -0.1, size = 3,
           colour = "#B71C1C", fontface = "bold") +

  scale_colour_manual(values = agent_cols, guide = "none") +
  scale_x_datetime(date_labels = "%b %d", date_breaks = "3 days") +
  scale_y_discrete(expand = expansion(add = 0.8)) +

  labs(
    title    = "Panel 1 — Risk creators",
    subtitle = "Agents posting personally before the breach (triangles = posts)",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x   = element_text(angle = 45, hjust = 1),
    axis.text.y   = element_text(size = 11, face = "bold"),
    plot.title    = element_text(size = 12, face = "bold", colour = "#B71C1C"),
    plot.subtitle = element_text(size = 9, colour = "grey40"),
    panel.grid.minor = element_blank()
  )

# ── Panel 2: Risk FLAGGERS — agents raising warnings in deliberating ──────────
p_flaggers <- ggplot() +

  annotate("rect",
           xmin = as.numeric(ymd_hms("2046-05-17T00:00:00")),
           xmax = as.numeric(embargo_date),
           ymin = -Inf, ymax = Inf,
           fill = "#E3F2FD", alpha = 0.4) +
  annotate("rect",
           xmin = as.numeric(embargo_date),
           xmax = as.numeric(breach_time),
           ymin = -Inf, ymax = Inf,
           fill = "#FFF3E0", alpha = 0.4) +

  # warning signal points
  geom_point(
    data  = pre_warnings,
    aes(x = timestamp, y = agent, colour = agent),
    shape = 18, size = 5, alpha = 0.85
  ) +

  # risk score overlay (Platform Trust only)
  geom_point(
    data   = pre_risk_scores,
    aes(x  = timestamp, y = agent),
    colour = "#1D9E75", shape = 16, size = 4
  ) +
  geom_label(
    data  = pre_risk_scores,
    aes(x = timestamp, y = agent,
        label = paste0("Risk: ", risk_score, "/10")),
    size  = 3.5, nudge_y = 0.4,
    colour = "#1D9E75",
    label.padding = unit(0.15, "lines"),
    fill  = "white", show.legend = FALSE
  ) +

  # warning excerpt labels
  geom_text_repel(
    data  = pre_warnings %>%
              mutate(short = str_trunc(excerpt, 40)),
    aes(x = timestamp, y = agent,
        label = short, colour = agent),
    size         = 3.2,
    max.overlaps = 10,
    box.padding  = 0.5,
    show.legend  = FALSE
  ) +

  # embargo and breach lines
  geom_vline(xintercept = as.numeric(embargo_date),
             colour = "#E65100", linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = embargo_date,
           y = length(unique(pre_warnings$agent)) + 0.5,
           label = "Embargo (May 23)",
           hjust = -0.1, size = 4,
           colour = "#E65100", fontface = "italic") +

  geom_vline(xintercept = as.numeric(breach_time),
             colour = "#B71C1C", linewidth = 1.2) +
  annotate("text", x = breach_time,
           y = length(unique(pre_warnings$agent)) + 0.5,
           label = "@Elena (May 29)",
           hjust = -0.1, size = 3,
           colour = "#B71C1C", fontface = "bold") +

  scale_colour_manual(values = agent_cols, guide = "none") +
  scale_x_datetime(date_labels = "%b %d", date_breaks = "3 days") +
  scale_y_discrete(expand = expansion(add = 0.8)) +

  labs(
    title    = "Panel 2 — Risk flaggers",
    subtitle = "Agents raising warnings in private deliberating (diamonds = warnings, circles = risk score)",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x   = element_text(angle = 45, hjust = 1, size=11),
    axis.text.y   = element_text(size = 12, face = "bold"),
    plot.title    = element_text(size = 14, face = "bold", colour = "#1D9E75"),
    plot.subtitle = element_text(size = 11, colour = "grey40"),
    panel.grid.minor = element_blank()
  )

# ── combine ───────────────────────────────────────────────────────────────────
(p_creators / p_flaggers) +
  plot_annotation(
    title    = "Leading indicators — who created risk vs who flagged it",
    subtitle = paste0(
      "Top: agents whose personal posting behaviour created the conditions for the breach\n",
      "Bottom: agents who privately flagged growing risk but took no formal action"
    ),
    theme = theme(
      plot.title    = element_text(size = 16, face = "bold"),
      plot.subtitle = element_text(size = 12, colour = "grey40")
    )
  )
```

::::

## 6 Conclusion

The investigation found no indication of a deliberate plan to bypass regulations. Rather it exposes a governance framework that fails to anticipate and prevent a foreseeable breach. Risk indicators shot up for nearly two weeks, warning signs piled up across several actors, and behavioural patterns appeared well before the disclosure. But these signals were segregated in discrete channels of communication, with no real escalation or enforcement mechanism. By the time @Elena’s post arrived on May 29, the system was already dysfunctional. The breach was not the result of a single mistake but the inevitable consequence of insufficient controls, informal processes and delayed oversight. The primary question for CivicLoom is whether such a governance architecture was up to the task of protecting a sensitive merger transaction. Based on the evidence that we have at this point, the answer is an unequivocal 'No'.